---
title: "Homework 4"
output: 
  rmdformats::robobook:
    css: "homework.css"
    highlight: pygments
link-citations: yes
editor_options: 
  chunk_output_type: console
---

**Due:** Friday, September 26, 10:00pm on Canvas

**Instructions:** 

**Code guidelines:**

* If a question requires code, and code is not provided, you will not receive full credit
* You will be graded on the quality of your code. In addition to being correct, your code should also be easy to read

## Conceptual questions

### Cook's distance

Below is the output for a logistic regression model, and the Cook's distance plot for the fitted model.

```{r, include=F}
set.seed(42)

x <- rnorm(100)

# create P(Y = 1 | X) for each entry in x
# Here log odds = -1 + 2x
p <- exp(-1 + 2*x)/(1 + exp(-1 + 2*x))

# Finally, simulate a binary response at each x
y <- rbinom(100, 1, p)

x <- c(x, 2.5, 3)
y <- c(y, 0, 1)

m1 <- glm(y ~ x, family = binomial)
```

```{r, echo=F}
summary(m1)
```

```{r, echo=F, message=F, warning=F}
library(tidyverse)
library(car)
library(mvtnorm)

data.frame(x = x, y = cooks.distance(m1)) |>
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  theme_bw() +
  labs(x = "x", y = "Cook's distance")
```


:::{.question}
#### Question 1

The largest Cook's distance is for an observation with $X = 2.5$. However, there is a point with $X = 3$ which has a smaller Cook's distance. How has this happened? In your answer, use the output provided to discuss what the observed value of $Y$ is for each of these two points.
:::

### Variance inflation factors

Consider a logistic regression model with systematic component

$$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + \cdots + \beta_k X_{ik}$$

To check for multicollinearity in the explanatory variables (and assess its impact on the estimated coefficients of the model), recall that we can calculate a *variance inflation factor*. The VIF for the $j$th coefficient is defined by

$$VIF_j = \dfrac{Var(\widehat{\beta}_j) \ \text{in the full model (other variables included)}}{Var(\widehat{\beta}_j) \ \text{when } X_j \text{ is the only variable in the model}}$$

Suppose we have three explanatory variables, $X_1$, $X_2$, and $X_3$. We fit a logistic regression model with only $X_1$; here are the results:

```{r, include=F}
set.seed(383)

n <- 1000
rho <- 0.98
beta <- c(0, 1, 1, 1)

# Create the covariance matrix sigma
sigma <- diag(rep(1 - rho, 3)) + matrix(rep(rho, 9), nrow=3)

# Generate Xs from a multivariate normal distribution
X <- rmvnorm(n, sigma = sigma)

x1 <- X[,1]
x2 <- X[,2]
x3 <- X[,3]

X_complete <- cbind(1, X)

# generate probabilities p
p <- exp(X_complete %*% beta)/(1 + exp(X_complete %*% beta))

# generate bernoulli response y
y <- rbinom(n, 1, p)
```

```{r}
m1 <- glm(y ~ x1, family = binomial)

summary(m1)
```

From the output, notice that $SE(\widehat{\beta}_1) \approx 0.212$.

We also fit a model with all three variables, and calculate the VIFs:

```{r}
m2 <- glm(y ~ x1 + x2 + x3, family = binomial)
vif(m2)
```

:::{.question}
#### Question 2

Using the output provided for the two models, calculate $SE(\widehat{\beta}_1)$ for the *second* model, `m2`.
:::

### Deviance

Below is the output for a logistic regression model:


```{r, include=F}
set.seed(383)

n <- 100
rho <- 0.99
beta <- c(0, 2, 1, 1)

# Create the covariance matrix sigma
sigma <- diag(rep(1 - rho, 3)) + matrix(rep(rho, 9), nrow=3)

# Generate Xs from a multivariate normal distribution
X <- rmvnorm(n, sigma = sigma)

x1 <- X[,1]
x2 <- X[,2]
x3 <- X[,3]

X_complete <- cbind(1, X)

# generate probabilities p
p <- exp(X_complete %*% beta)/(1 + exp(X_complete %*% beta))

# generate bernoulli response y
y <- rbinom(n, 1, p)
```

```{r}
m1 <- glm(y ~ x1, family = binomial)

summary(m1)
```


:::{.question}
#### Question 3

What is the log-likelihood for the fitted model?
:::

:::{.question}
#### Question 4

The estimated coefficients for the model are $\widehat{\beta}_0 = -0.4769$ and $\widehat{\beta}_1 = 3.7934$.

Suppose that instead of these estimated coefficients, I decide I prefer $\widehat{\beta}_0 = -0.5$ and $\widehat{\beta}_1 = 4$ (on the same data). If I calculate the log-likelihood and deviance for the model with these different coefficients, how would they compare to the output shown above? (For each of  log-likelihood and deviance, explain whether it will increase, decrease, or stay the same).
:::

We now fit a second model, with two additional explanatory variables:

```{r}
m2 <- glm(y ~ x1 + x2 + x3, family = binomial)
summary(m2)
vif(m2)
```

Comparing the two models (`m1` and `m2` in the displayed output), we notice that while the deviance decreases very slightly, the AIC has *increased* in model 2. 

:::{.question}
#### Question 5

Explain what is causing the AIC to increase when we add additional variables to the model. Your answer should use information provided by the displayed VIFs.
:::


### Prediction

Your friend has built a logistic regression model to diagnose a rare disease, using a sample of 10000 patients from providers across NC. To assess their model's predictive ability, they create a confusion matrix:

| | Disease negative | Disease positive|
| --- | --- | --- |
|Prediction negative | 9501 | 489 |
| Prediction positive | 6 | 4 |

They calculate the overall accuracy of their predictions, and are ecstatic to obtain an accuracy of 95%!

:::{.question}
#### Question 6

Should your friend be happy with their model's predictive performance? Explain why or why not. Your answer should include a discussion of other metrics for assessing predictive ability (such as sensitivity and specificity).
:::


### More prediction

A different friend of yours is feeling ill, and goes to their doctor. Their doctor diagnoses them with a disease called the *Red Death* (what a horrifying name!). Understandably upset, your friend asks about the reliability of the diagnostic test.

In response, their doctor provides a confusion matrix from a recent study on the Red Death:

| | Disease negative | Disease positive|
| --- | --- | --- |
|Test negative | 8003 | 10 |
| Test positive | 312 | 95 |

The doctor claims that, *if a patient is positive for the disease*, there is a 90.5% chance that they will test positive. Therefore, the doctor argues that your friend is very likely to truly have the disease.

:::{.question}
#### Question 7

Using the confusion matrix, explain how the doctor got the 90.5% probability they reported.
:::

:::{.question}
#### Question 8

*Given that your friend tested positive*, how likely is it that they actually have the disease? Calculate a relevant value from the confusion matrix. 

Is the doctor correct that your friend is "very likely to truly have the disease"?
:::








