---
title: Prediction
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Diagnostics summary

.large[
Before fitting a model:

* Check empirical logit plots; make transformations as needed
* Check for multicollinearity with scatterplots / correlations between quantitative explanatory variables; if needed, choose a subset of variables or combine variables

After fitting a model:

* Check quantile residual plots
* Check for multicollinearity with VIFs
* Check for influential points with Cook's distance
    * If influential points are clearly an error, can remove
    * Otherwise, report results with and without influential points
]

---

### Using logistic regression

.large[
Class so far:

* Model fitting
* Model interpretation
* Calculating predictions
* Assessing model assumptions

What's next:

* Assessing predictions
* Variable selection
* Inference
]

---

### Warmup

.large[
Work on the warmup activity (handout), then we will discuss as a group
]

---

### Class activity

.large[
Fitted model:
]

$\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = 3.78 - 0.037 Age_i - 2.52 Male_i - 1.31 Class2_i - 2.58 Class3_i$

.large[
.question[
What is the predicted probability of survival for a male, first-class passenger aged 20? What about for a male, second-class passenger aged 30?
]
]

---

### Making predictions with the Titanic data

.large[
* For each passenger, we calculate $\widehat{\pi}_i$ (estimated probability of survival)
* But, we want to predict *which* passengers actually survive

.question[
How do we turn $\widehat{\pi}_i$ into a binary prediction of survival / no survival?
]
]

---

### Confusion matrix

```{r, include=F}
library(tidyverse)
titanic <- read.csv("https://sta214-s23.github.io/homework/Titanic.csv")
```

.large[
```{r, eval=F}
m1 <- glm(Survived ~ Age + Sex + as.factor(Pclass), 
          data = titanic, family = binomial)

table(Prediction = m1$fitted.values > 0.5, 
      Truth = titanic$Survived)
```

```{r, echo=F}
m1 <- glm(Survived ~ Age + Sex + as.factor(Pclass), 
          data = titanic, family = binomial)

table(Prediction = m1$fitted.values > 0.5, 
      Truth = m1$y)
```

.question[
Did we do a good job predicting survival?
]
]

---

### Confusion matrix

.large[
```{r, echo=F}
table(Prediction = m1$fitted.values > 0.5, 
      Truth = m1$y)
```
]

---

### Class activity

.large[
Predictions with the SBA data:

[https://sta214-f25.github.io/class_activities/ca_10.html](https://sta214-f25.github.io/class_activities/ca_10.html)
]


---

### Class activity

```{r, include=F}
sba <- read.csv("https://sta214-s23.github.io/class_activities/sba_small.csv")
```

.large[
```{r}
m1 <- glm(Default ~ log(DisbursementGross) + Term + 
            as.factor(UrbanRural), 
          family = binomial, data = sba)

table(Prediction = m1$fitted.values > 0.5, 
      Truth = sba$Default)
```

* Accuracy = 
* Sensitivity = 
* Specificity = 
* PPV = 
]

---

### Class activity

.large[
```{r, echo=F}
table(Prediction = m1$fitted.values > 0.5, 
      Truth = sba$Default)
```

.question[
Is an accuracy of around 80% good?
]
]

---

### Class activity

.large[
Changing thresholds:

```{r}
table(Prediction = m1$fitted.values > 0.3, 
      Truth = sba$Default)
```

```{r}
table(Prediction = m1$fitted.values > 0.7, 
      Truth = sba$Default)
```

]