---
title: Logistic regression assumptions and diagnostics
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Agenda and reminders

.large[
* HW 2 due Friday 
* Department seminar tomorrow (9/11) at 11am in ZSR auditorium
    * please refrain from wearing colognes, perfumes, and/or heavily scented body and hair products
* Today: more on diagnostics
]

---

### Shape assumption and quantile residuals

.large[
(In R)
]

---

### Multicollinearity

.large[
.question[
What is multicollinearity?
]
]

---

### Example: College scorecard data

.large[
The Department of Education compiles the College Scorecard, which is used to help prospective college students compare schools. For each school, variables include:

* CONTROL: whether the school is public or private
* SATVRMID: midpoint of SAT critical reading scores of students attending the school
* ACTCMMID: midpoint of the ACT cumulative scores
* UGDS: number of undergraduate students at the school
* NPT4: average cost to attend the school
* PCTFLOAN: fraction of undergraduates receiving a federal student loan
* MD_EARN_WNE_P10: median salary of students 10 years after graduation
]

---

### Example: College scorecard data

.large[

* CONTROL: whether the school is public or private
* SATVRMID: midpoint of SAT critical reading scores of students attending the school
* ACTCMMID: midpoint of the ACT cumulative scores
* UGDS: number of undergraduate students at the school
* NPT4: average cost to attend the school
* PCTFLOAN: fraction of undergraduates receiving a federal student loan
* MD_EARN_WNE_P10: median salary of students 10 years after graduation

.question[
Which of these variables may suffer from multicollinearity?
]
]

---

### The impact of multicollinearity

.large[
How does correlation between the explanatory variables impact the fitted model?

* Increased variability in coefficient estimates
* Difficulty interpreting coefficients
]


---

### Diagnosing multicollinearity

.large[
.question[
How do I detect multicollinearity in my data?
]
]

---

### Scatterplot and correlation matrix

```{r, include=F}
library(tidyverse)
library(GGally)
scorecard <- read_csv("https://sta112-s22.github.io/projects/scorecard.csv")
```

```{r, message=F, warning=F, echo=F, fig.align='center', fig.width=10, fig.height=7}
scorecard %>%
  ggpairs(columns = 4:10,
          upper = list(continuous = "points"),
          lower = list(continuous = GGally::wrap(ggally_cor, stars = F))) +
  theme_classic()
```

---

### Variance inflation factors

---

### Variance inflation factors in R

```{r, message=F, warning=F}
library(car)

m1 <- glm(RPY_3YR_70 ~ CONTROL + SATVRMID + 
            SATMTMID + SATWRMID + ACTCMMID + ACTENMID + 
            ACTMTMID + ACTWRMID + UGDS +
            PCTFLOAN + MD_EARN_WNE_P10 + NPT4,
          data = scorecard, family = binomial)
vif(m1)
```

---

### Addressing model issues

.large[
.question[
How should we handle multicollinearity in the explanatory variables? Discuss with a neighbor for a few minutes, then we will discuss as a group.
]
]

---

### Outliers and influential points

.large[
.question[
What is an outlier?
]
]

---

### Cook's distance

---

### Addressing model issues

.large[
.question[
How should we handle outliers and influential points? Discuss with a neighbor for a few minutes, then we will discuss as a group.
]
]

---

### Class activity

.large[
[https://sta214-f25.github.io/class_activities/ca_08.html](https://sta214-f25.github.io/class_activities/ca_08.html)

.question[
At the end of class, render to HTML and submit HTML on Canvas
]
]