---
title: Linear mixed effects models
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=F}
library(lme4)
library(tidyverse)

bnb <- read.csv("https://raw.githubusercontent.com/proback/BYSH/master/data/airbnb.csv")
```


### Warmup

.large[
Work on the warmup activity (handout).
]

---

### Warmup

.large[
.question[
What are the groups in the data?
]
]

---

### Warmup

.large[
.question[
What mixed effects model should the researchers use to investigate their hypothesis?
]
]

---

### Assumptions

.large[
$$Price_{ij} = \beta_0 + \beta_1 Satisfaction_{ij} + u_i + \varepsilon_{ij}$$
$u_i \overset{iid}{\sim} N(0, \sigma_u^2) \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$
]

.large[
* **Shape:** 
  * the overall relationship between satisfaction and price is linear
  * The slope is the *same* for each neighborhood
]

---

### Assumptions

.large[
$$Price_{ij} = \beta_0 + \beta_1 Satisfaction_{ij} + u_i + \varepsilon_{ij}$$
$u_i \overset{iid}{\sim} N(0, \sigma_u^2) \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$
]

.large[
* **Independence:** 
  * random effects are independent
  * observations within neighborhoods are independent after accounting for the random effect (i.e., the random effect captures the correlation within neighborhoods)
  * observations from different neighborhoods are independent
]

---

### Assumptions

.large[
$$Price_{ij} = \beta_0 + \beta_1 Satisfaction_{ij} + u_i + \varepsilon_{ij}$$
$u_i \overset{iid}{\sim} N(0, \sigma_u^2) \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$
]

.large[
* **Normality:** Both $u_i \sim N(0, \sigma_u^2)$ and $\varepsilon_{ij} \sim N(0, \sigma_\varepsilon^2)$
* **Constant variance:** 
  * $\varepsilon_{ij}$ has the same variance $\sigma_\varepsilon^2$ regardless of satisfaction or neighborhood
]

---


### Calculating predicted values

.large[
Residuals: $Price_{ij} - \widehat{Price}_{ij}$, where

$$\widehat{Price}_{ij} = \widehat{\beta}_0 + \widehat{\beta}_1 Satisfaction_{ij} + \widehat{u}_i$$

.question[
Where do the $\widehat{u}_i$ come from?
]
]

---

### Estimated random effects

.large[
R calculates an estimated random effect for each group (i.e., neighborhood):

```{r}
m1 <- lmer(price ~ overall_satisfaction + 
             (1 | neighborhood), 
           data = bnb)
coef(m1)
```
]

---

### Estimated random effects

.large[

```{r, eval=F}
coef(m1)
```

```
                   (Intercept) overall_satisfaction
Albany Park          16.367331             14.80912
Archer Heights        9.863461             14.80912
Avondale             40.533851             14.80912
Beverly              21.046464             14.80912
Bridgeport           13.304517             14.80912
```

.question[
What is the same for every neighborhood?
]
]

---

### Estimated random effects

.large[

```{r, eval=F}
coef(m1)
```

```
                   (Intercept) overall_satisfaction
Albany Park          16.367331             14.80912
Archer Heights        9.863461             14.80912
Avondale             40.533851             14.80912
Beverly              21.046464             14.80912
Bridgeport           13.304517             14.80912
```

.question[
What is the same for every neighborhood?
]
]

.large[
The slope ( $\widehat{\beta}_1$ )
]

---

### Estimated random effects

.large[

```{r, eval=F}
coef(m1)
```

```
                   (Intercept) overall_satisfaction
Albany Park          16.367331             14.80912
Archer Heights        9.863461             14.80912
Avondale             40.533851             14.80912
Beverly              21.046464             14.80912
Bridgeport           13.304517             14.80912
```

.question[
What is *different* for each neighborhood?
]
]



---

### Estimated random effects

.large[

```{r, eval=F}
coef(m1)
```

```
                   (Intercept) overall_satisfaction
Albany Park          16.367331             14.80912
Archer Heights        9.863461             14.80912
Avondale             40.533851             14.80912
Beverly              21.046464             14.80912
Bridgeport           13.304517             14.80912
```

.question[
What is *different* for each neighborhood?
]
]

.large[
The intercept ( $\widehat{\beta}_0 + \widehat{u}_i$ )
]

---

### Estimated random effects

.large[

```
                   (Intercept) overall_satisfaction
Albany Park          16.367331             14.80912
Archer Heights        9.863461             14.80912
Avondale             40.533851             14.80912
Beverly              21.046464             14.80912
Bridgeport           13.304517             14.80912
```

.question[
How do I get the random effect estimates $\widehat{u}_i$ ?
]
]

---

### Estimated random effects

.large[

```
                   (Intercept) overall_satisfaction
Albany Park          16.367331             14.80912
Archer Heights        9.863461             14.80912
Avondale             40.533851             14.80912
Beverly              21.046464             14.80912
Bridgeport           13.304517             14.80912
```

.question[
How do I get the random effect estimates $\widehat{u}_i$ ?
]
]

.large[
The intercept for each neighborhood is $\widehat{\beta}_0 + \widehat{u}_i$

So subtract the estimate $\widehat{\beta}_0$ from the intercept for each neighborhood
]

---

### Checking assumptions

.large[
$$Price_{ij} = \beta_0 + \beta_1 Satisfaction_{ij} + u_i + \varepsilon_{ij}$$
$u_i \overset{iid}{\sim} N(0, \sigma_u^2) \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$
]

.large[
* **Normality assumption:** Both $u_i \sim N(0, \sigma_u^2)$ and $\varepsilon_{ij} \sim N(0, \sigma_\varepsilon^2)$
  
.question[
How do you think we could check the normality assumption?
]
]

---

### QQ plots

.large[
**Assumption:** $u_i \sim N(0, \sigma_u^2)$

* Check whether the random effect estimates $\widehat{u}_i$ appear normal with a QQ plot

**Assumption:** $\varepsilon_{ij} \sim N(0, \sigma_\varepsilon^2)$

* Check whether the residuals appear normal with a QQ plot
]

---

### QQ plot for the residuals

```{r message=F, warning=F, echo=F, fig.align='center', fig.height=5, fig.width=7}

bnb %>%
  mutate(resids = residuals(m1)) %>%
  ggplot(aes(sample = resids)) +
  geom_qq(size=2) +
  geom_qq_line(color="blue", lwd=1.2) +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed residual quantiles") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
.question[
Do the residuals appear normal?
]
]

---

### QQ plot for the residuals

```{r message=F, warning=F, echo=F, fig.align='center', fig.height=5, fig.width=7}

bnb %>%
  mutate(resids = residuals(m1)) %>%
  ggplot(aes(sample = resids)) +
  geom_qq(size=2) +
  geom_qq_line(color="blue", lwd=1.2) +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed residual quantiles") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
.question[
Do the residuals appear normal?
]
]

.large[
No -- though the sample size is large, so it might not matter
]

---

### QQ plot for the random effects

```{r message=F, warning=F, echo=F, fig.align='center', fig.height=5, fig.width=7}

data.frame(re = coef(m1)$neighborhood[,1] - summary(m1)$coefficients[2,1]) %>%
  ggplot(aes(sample = re)) +
  geom_qq(size=2) +
  geom_qq_line(color="blue", lwd=1.2) +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed random effect quantiles") +
  theme_bw() +
  theme(text = element_text(size = 20))
```

.large[
.question[
Do the random effects appear normal?
]
]

---

### QQ plot for the random effects

```{r message=F, warning=F, echo=F, fig.align='center', fig.height=5, fig.width=7}

data.frame(re = coef(m1)$neighborhood[,1] - summary(m1)$coefficients[2,1]) %>%
  ggplot(aes(sample = re)) +
  geom_qq(size=2) +
  geom_qq_line(color="blue", lwd=1.2) +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed random effect quantiles") +
  theme_bw() +
  theme(text = element_text(size = 20))
```

.large[
.question[
Do the random effects appear normal?
]
]

.large[
No -- and this could be more problematic
]

---

### Addressing assumption violations


.pull-left[
```{r message=F, warning=F, echo=F, fig.align='center', fig.height=4, fig.width=6}

bnb %>%
  mutate(resids = residuals(m1)) %>%
  ggplot(aes(sample = resids)) +
  geom_qq(size=2) +
  geom_qq_line(color="blue", lwd=1.2) +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed residual quantiles") +
  theme_bw() +
  theme(text = element_text(size = 20))
```
]

.pull-right[
```{r message=F, warning=F, echo=F, fig.align='center', fig.height=4, fig.width=6}

data.frame(re = coef(m1)$neighborhood[,1] - summary(m1)$coefficients[2,1]) %>%
  ggplot(aes(sample = re)) +
  geom_qq(size=2) +
  geom_qq_line(color="blue", lwd=1.2) +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed random effect quantiles") +
  theme_bw() +
  theme(text = element_text(size = 20))
```
]

.large[
.question[
How could we address violations of the normality assumptions?
]
]

---

### Addressing assumption violations


.pull-left[
```{r message=F, warning=F, echo=F, fig.align='center', fig.height=4, fig.width=6}

bnb %>%
  mutate(resids = residuals(m1)) %>%
  ggplot(aes(sample = resids)) +
  geom_qq(size=2) +
  geom_qq_line(color="blue", lwd=1.2) +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed residual quantiles") +
  theme_bw() +
  theme(text = element_text(size = 20))
```
]

.pull-right[
```{r message=F, warning=F, echo=F, fig.align='center', fig.height=4, fig.width=6}

data.frame(re = coef(m1)$neighborhood[,1] - summary(m1)$coefficients[2,1]) %>%
  ggplot(aes(sample = re)) +
  geom_qq(size=2) +
  geom_qq_line(color="blue", lwd=1.2) +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed random effect quantiles") +
  theme_bw() +
  theme(text = element_text(size = 20))
```
]

.large[
.question[
How could we address violations of the normality assumptions?
]
]

.large[
Transformations!
]

---

### Transformations

.large[
$$\log(Price_{ij}) = \beta_0 + \beta_1 Satisfaction_{ij} + u_i + \varepsilon_{ij}$$
$u_i \overset{iid}{\sim} N(0, \sigma_u^2) \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$
]

```{r include =F}
m1 <- lmer(log(price) ~ overall_satisfaction + (1 | neighborhood), data = bnb)
```

.pull-left[
```{r message=F, warning=F, echo=F, fig.align='center', fig.height=4, fig.width=6}

bnb %>%
  mutate(resids = residuals(m1)) %>%
  ggplot(aes(sample = resids)) +
  geom_qq(size=2) +
  geom_qq_line(color="blue", lwd=1.2) +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed residual quantiles") +
  theme_bw() +
  theme(text = element_text(size = 20))
```
]

.pull-right[
```{r message=F, warning=F, echo=F, fig.align='center', fig.height=4, fig.width=6}

data.frame(re = coef(m1)$neighborhood[,1] - summary(m1)$coefficients[2,1]) %>%
  ggplot(aes(sample = re)) +
  geom_qq(size=2) +
  geom_qq_line(color="blue", lwd=1.2) +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed random effect quantiles") +
  theme_bw() +
  theme(text = element_text(size = 20))
```
]

.large[
* Transformations can help with both the residuals and random effects
]

---

### Class activity

.large[
Work on the class activity on the course website. Submit your HTML on Canvas when finished.
]





