---
author: "Ciaran Evans"
title: Logistic regression interpretation
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r include=F}
dengue <- read.csv("https://sta279-s22.github.io/labs/dengue.csv")
```


### Recap: logistic regression model

.large[

$Y_i =$ dengue status (0 = negative, 1 = positive) 

$Age_i =$ age (in years)

**Random component:** $\hspace{1cm} Y_i \sim Bernoulli(\pi_i)$

**Systematic component:** $\hspace{1cm} \log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 \ Age_i$

]

---

### Recap: interpretation

.large[
$$\hspace{1cm} \log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = -2.45 + 0.22 \ Age_i$$

**Slope:**

* A one-year increase in age is associated with an increase in the *log odds* of dengue of 0.22.
* A one-year increase in age is associated with an increase in the *odds* of dengue by a *factor* of $e^{0.22} = 1.25$.

**Intercept:**

* when age is 0, the estimated log odds of dengue is $-2.45$
* when age is 0, the estimated odds of dengue is $e^{-2.45} = 0.086$
]

---

### Recap: adding more variables

.large[
Now let's add WBC as a variable to the model:

```{r, eval=F}
m2 <- glm(Dengue ~ Age + WBC, data = dengue,
          family = binomial)
summary(m2)
```

$$\hspace{1cm} \log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = 0.34 + 0.15 \ Age_i - 0.31 WBC_i$$

A one-year increase in age is associated with an increase in the odds of dengue by a factor of $e^{0.15} = 1.16$, holding white blood cell count fixed.
]

---

### Adding more variables

.large[
What about a *categorical* predictor?

```{r, eval=F}
m3 <- glm(Dengue ~ Age + Abdominal, data = dengue,
          family = binomial)
summary(m3)
```

$$\hspace{1cm} \log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = -0.25 + 0.22 \ Age_i + 0.21 Abdominal_i$$

.question[
How should I interpret the coefficient on abdominal pain?
]
]

---

### Interpretation

.large[
$$\hspace{1cm} \log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = -2.45 + 0.22 \ Age_i$$

**Slope:**

* A one-year increase in age is associated with an increase in the *log odds* of dengue of 0.22.
* A one-year increase in age is associated with an increase in the *odds* of dengue by a *factor* of $e^{0.22} = 1.25$.

.question[
Can I interpret the coefficient in terms of the *probability* (rather than odds or log odds)?
]
]


