---
title: Inference and linear combinations
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Last time: Activity

.large[
```
Coefficients:
               Estimate Std. Error z value Pr(>|z|)    
(Intercept)   -3.989979   1.139951  -3.500 0.000465 ***
gre            0.002264   0.001094   2.070 0.038465 *  
gpa            0.804038   0.331819   2.423 0.015388 *  
factor(rank)2 -0.675443   0.316490  -2.134 0.032829 *  
factor(rank)3 -1.340204   0.345306  -3.881 0.000104 ***
factor(rank)4 -1.551464   0.417832  -3.713 0.000205 ***
```

.question[
Given your friend's GPA (3.95), school (rank 3), and current GRE score (590), what is your estimated probability that they will be accepted to graduate school?
]
]

---

### Activity

.large[
```
Coefficients:
               Estimate Std. Error z value Pr(>|z|)    
(Intercept)   -3.989979   1.139951  -3.500 0.000465 ***
gre            0.002264   0.001094   2.070 0.038465 *  
gpa            0.804038   0.331819   2.423 0.015388 *  
factor(rank)2 -0.675443   0.316490  -2.134 0.032829 *  
factor(rank)3 -1.340204   0.345306  -3.881 0.000104 ***
factor(rank)4 -1.551464   0.417832  -3.713 0.000205 ***
```

.question[
The standard deviation of GRE scores is 115.5 points. If your friend re-took the GRE and increased their score by one standard deviation, what would be the estimated change in their *odds* of being accepted?
]
]

---

### Activity

.large[
```
Coefficients:
               Estimate Std. Error z value Pr(>|z|)    
(Intercept)   -3.989979   1.139951  -3.500 0.000465 ***
gre            0.002264   0.001094   2.070 0.038465 *  
gpa            0.804038   0.331819   2.423 0.015388 *  
factor(rank)2 -0.675443   0.316490  -2.134 0.032829 *  
factor(rank)3 -1.340204   0.345306  -3.881 0.000104 ***
factor(rank)4 -1.551464   0.417832  -3.713 0.000205 ***
```

.question[
If your friend re-took the GRE and increased their score by one standard deviation, what would be their new estimated *probability* of getting accepted?
]
]

---

### Activity

.large[
```
Coefficients:
               Estimate Std. Error z value Pr(>|z|)    
(Intercept)   -3.989979   1.139951  -3.500 0.000465 ***
gre            0.002264   0.001094   2.070 0.038465 *  
gpa            0.804038   0.331819   2.423 0.015388 *  
factor(rank)2 -0.675443   0.316490  -2.134 0.032829 *  
factor(rank)3 -1.340204   0.345306  -3.881 0.000104 ***
factor(rank)4 -1.551464   0.417832  -3.713 0.000205 ***
```

.question[
Calculate a 95% confidence interval for the change in *odds* of acceptance associated with a *one-unit* increase in GRE score, holding GPA and school rank fixed
]
]

---

### Activity

.large[
```
Coefficients:
               Estimate Std. Error z value Pr(>|z|)    
(Intercept)   -3.989979   1.139951  -3.500 0.000465 ***
gre            0.002264   0.001094   2.070 0.038465 *  
gpa            0.804038   0.331819   2.423 0.015388 *  
factor(rank)2 -0.675443   0.316490  -2.134 0.032829 *  
factor(rank)3 -1.340204   0.345306  -3.881 0.000104 ***
factor(rank)4 -1.551464   0.417832  -3.713 0.000205 ***
```

.question[
Now calculate a 95% confidence interval for the change in *odds* of acceptance associated with a one *standard deviation* increase in GRE score, holding GPA and school rank fixed
]
]

---

### Activity

.large[
95% confidence interval for change in *odds* of acceptance associated with a one *standard deviation* increase in GRE score, holding GPA and school rank fixed:

$$(1.014, 1.66)$$

Your friend is reluctant to retake the exam unless they believe that their odds of acceptance will increase by at least 25%.

.question[
Would you recommend that your friend re-takes the exam?
]
]

---

### A new question

.large[
So far:

* Inference for the *change* in log-odds ($\beta_1$) or odds ($e^{\beta_1}$) for a one-unit increase in $X$
* Inference for change in the log-odds and odds for a one *standard deviation* increase in $X$

New question:

.question[
Can we do inference about a *probability*? E.g., can we construct a confidence interval for the true probability of admission, given values of the explanatory variables?
]
]

---

### Example

.large[
For simplicity, suppose we just look at GRE score for now:

```
             Estimate Std. Error z value Pr(>|z|)    
(Intercept) -2.901344   0.606038  -4.787 1.69e-06 ***
gre          0.003582   0.000986   3.633  0.00028 ***
```

For a student with a GRE of 590, the *estimated* probability of admission is $\widehat{\pi}_i = 0.313$.

.question[
We want to construct a confidence interval for the *true* probability of admission, $\pi_i$, for this student.
]
]

---

### Example

.large[
For simplicity, suppose we just look at GRE score for now:

```
             Estimate Std. Error z value Pr(>|z|)    
(Intercept) -2.901344   0.606038  -4.787 1.69e-06 ***
gre          0.003582   0.000986   3.633  0.00028 ***
```

For a student with a GRE of 590, the *estimated* probability of admission is $\widehat{\pi}_i = 0.313$.

.question[
We want to construct a confidence interval for the *true* probability of admission, $\pi_i$, for this student.
]

* create a confidence interval for the true *log odds* of admission for this student
* transform the interval into probabilities
]

---

### Confidence interval for the log odds

.large[
For a student with a GRE of 590, the *estimated* log odds is:

$$\widehat{\text{log odds}} = \widehat{\beta}_0 + \widehat{\beta}_1 (590)$$

A confidence interval for the *true* log odds is:

$$\widehat{\text{log odds}} \pm z^* SE(\widehat{\text{log odds}})$$
]


---

### Variance of a sum of random variables

.large[
**Fact:** For two variables $U$, $V$ and constants $a, b$:

$$Var(aU + bV) = a^2 Var(U) + b^2 Var(V) + 2ab Cov(U, V)$$

where $Cov(U, V) = \text{Correlation}(U, V) \cdot SD(U) \cdot SD(V)$
]

---

### Finding the variance and covariance

.large[
In R, `vcov` can be used to extract variances and covariances for estimated coefficients:

```{r, include=F}
library(tidyverse)
grad_app <- read.csv("https://stats.idre.ucla.edu/stat/data/binary.csv")
```


```{r}
m1 <- glm(admit ~ gre, family = binomial, 
          data = grad_app)
vcov(m1)
```

]

---

### Confidence interval for the log odds

.large[
For a student with GRE 590, $SE(\widehat{\text{log odds}}) = 0.11$

95% confidence interval for log odds:

$$\widehat{\text{log odds}} \pm 1.96 \ SE(\widehat{\text{log odds}})$$
]

---

### Confidence interval for the probability

.large[
95% confidence interval for the log odds, for a student with GRE 590:

$$(-1.004, -0.572)$$

.question[
How would I turn this into a 95% confidence interval for the true *probability* of admission to graduate school?
]
]

---

### Class activity

.large[
Work on the class activity (handout).
]







