---
title: Random slopes
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Logistics

.large[
* Project part 3 due last day of class
  * No additional analysis needed; just combining work from first two parts and adding introduction, conclusion, etc.
  * Work time in class on Wednesday and Friday
* Homework 9 (optional) due Wednesday
* Final exam: December 13, 2 - 5pm in Carswell 101 (our usual room)
  * Final exam will be cumulative, and will include mixed effects models
]

---

### Revisiting the shape assumption

.large[
$$Price_{ij} = \beta_0 + \beta_1 Satisfaction_{ij} + u_i + \varepsilon_{ij}$$
$u_i \overset{iid}{\sim} N(0, \sigma_u^2) \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$
]

.large[
* This model assumes that the slope is the same for each neighborhood

.question[
How can we change the model to allow the slope to be different in different neighborhoods?
]
]

---

### Adding random slopes

.large[
$$Price_{ij} = \beta_0 + u_i + (\beta_1 + v_i) Satisfaction_{ij} + \varepsilon_{ij}$$

* $\beta_0$ = mean price when satisfaction is 0 (average across neighborhoods)
* $\beta_0 + u_i$ = mean price when satisfaction is 0 in neighborhood $i$
* $\beta_1$ = average change in price for a one-unit increase in satisfaction (average across neighborhoods)
* $\beta_1 + v_i$ = average change in price for a one-unit increase in satisfaction in neighborhood $i$
]

---

### Warmup

.large[
Work on the warmup activity (handout)
]

---

### Warmup

.large[
The researchers hypothesize that anxiety levels depend on the type of performance (large or small ensembles), and that the difference in anxiety levels between large and small ensembles varies from person to person.

.question[
What mixed effects model should the researchers use to investigate their hypothesis?
]
]

---

### Example: music performance anxiety

.large[
$$Anxiety_{ij} = \beta_0 + u_i + (\beta_1 + v_i) LargeEnsemble_{ij} + \varepsilon_{ij}$$

.question[
Interpret the fixed effects and random effects in the model.
]
]

---

### Example: music performance anxiety

.large[
$$Anxiety_{ij} = \beta_0 + u_i + (\beta_1 + v_i) LargeEnsemble_{ij} + \varepsilon_{ij}$$

.question[
Interpret the fixed effects and random effects in the model.
]
]

.large[
* $\beta_0$ = average performance anxiety before small ensemble and solo performances (average across musicians)
* $\beta_0 + u_i$ = average performance anxiety before small ensemble and solo performances for musician $i$
* $\beta_1$ = average difference in anxiety before large ensemble performances (compared to small/solo performances) (average across musicians)
* $\beta_1 + v_i$ = average difference in anxiety before large ensemble performances for musician $i$
]

---

### Fitting a model with random slopes

.large[
```{r, eval=F}
m1 <- lmer(na ~ large + (large|id),
           data = music)
```

* This means we include `large` as a fixed effect, *and* we allow the coefficient on `large` to vary from individual to individual
]

---

### Fitting a model with random slopes

```{r include=F}
library(tidyverse)
library(lme4)
library(knitr)

music <- read.csv("musicdata.csv") %>%
   mutate(large = ifelse(perform_type == "Large Ensemble", 1, 0))
```

.large[
```{r, eval=F}
m1 <- lmer(na ~ large + (large|id),
           data = music)
summary(m1)
```

```
Fixed effects:
            Estimate Std. Error t value
(Intercept)  16.7297     0.4908   34.09
large        -1.6762     0.5425   -3.09
```

$\widehat{\beta}_0 = 16.73 \hspace{1cm} \widehat{\beta}_1 = -1.68$

.question[
How would I interpret $\widehat{\beta}_0$ and $\widehat{\beta}_1$?
]
]

---

### Fitting a model with random slopes

.large[
$$Anxiety_{ij} = \beta_0 + u_i + (\beta_1 + v_i) LargeEnsemble_{ij} + \varepsilon_{ij}$$

$\widehat{\beta}_0 = 16.73 \hspace{1cm} \widehat{\beta}_1 = -1.68$

* The average anxiety before a small/solo performance is 16.73
* On average, student anxiety decreases by 1.68 before a large performance (compared to a small/solo performance)
]

---

### Fitting a model with random slopes

.large[
```{r, eval=F}
summary(m1)
```

```
Random effects:
 Groups   Name        Variance Std.Dev. Corr 
 id       (Intercept)  6.3330  2.5165        
          large        0.7429  0.8619   -0.76
 Residual             21.7712  4.6660        
```

.question[
What does this output tell us about the random effects and the noise?
]
]

---

### Fitting a model with random slopes


.large[
$$Anxiety_{ij} = \beta_0 + u_i + (\beta_1 + v_i) LargeEnsemble_{ij} + \varepsilon_{ij}$$

```
Random effects:
 Groups   Name        Variance Std.Dev. Corr 
 id       (Intercept)  6.3330  2.5165        
          large        0.7429  0.8619   -0.76
 Residual             21.7712  4.6660        
```

* $\widehat{\sigma}_u^2 = 6.333$ (variability in anxiety before small performances, between students)
* $\widehat{\sigma}_v^2 = 0.743$ (variability in difference in anxiety before large performances, between students)
* $\widehat{\sigma}_\varepsilon^2 = 21.77$ (variability in anxiety between performances, within a student)
]

---

### Correlation between slopes and intercepts

.large[
$$Anxiety_{ij} = \beta_0 + u_i + (\beta_1 + v_i) LargeEnsemble_{ij} + \varepsilon_{ij}$$

```
Random effects:
 Groups   Name        Variance Std.Dev. Corr 
 id       (Intercept)  6.3330  2.5165        
          large        0.7429  0.8619   -0.76
 Residual             21.7712  4.6660        
```

* $\widehat{\rho}_{uv} = -0.76$ (estimated correlation between the random slope and random intercept for an individual)
  * Subjects with higher performance anxiety scores for solos and small ensembles tend to have greater decreases in performance anxiety for large ensemble performances
]

---

### Writing down the model

.large[
$$Anxiety_{ij} = \beta_0 + u_i + (\beta_1 + v_i) LargeEnsemble_{ij} + \varepsilon_{ij}$$
$\varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2) \hspace{1cm} \begin{bmatrix} u_i \\ v_i \end{bmatrix} \overset{iid}{\sim} N \left( \begin{bmatrix} 0 \\ 0 \end{bmatrix}, \begin{bmatrix} \sigma_u^2 & \rho_{uv}\sigma_u \sigma_v \\ \rho_{uv}\sigma_u \sigma_v & \sigma_v^2 \end{bmatrix} \right)$

.question[
Anybody know the name of this new thing?
]
]

---

### Writing down the model

.large[
$$Anxiety_{ij} = \beta_0 + u_i + (\beta_1 + v_i) LargeEnsemble_{ij} + \varepsilon_{ij}$$
$\varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2) \hspace{1cm} \begin{bmatrix} u_i \\ v_i \end{bmatrix} \overset{iid}{\sim} N \left( \begin{bmatrix} 0 \\ 0 \end{bmatrix}, \begin{bmatrix} \sigma_u^2 & \rho_{uv}\sigma_u \sigma_v \\ \rho_{uv}\sigma_u \sigma_v & \sigma_v^2 \end{bmatrix} \right)$

.question[
Anybody know the name of this new thing?
]
]

.large[
A bivariate normal distribution!
]

---

### Writing down the model

.large[
$$Anxiety_{ij} = \beta_0 + u_i + (\beta_1 + v_i) LargeEnsemble_{ij} + \varepsilon_{ij}$$
$\varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2) \hspace{1cm} \begin{bmatrix} u_i \\ v_i \end{bmatrix} \overset{iid}{\sim} N \left( \begin{bmatrix} 0 \\ 0 \end{bmatrix}, \begin{bmatrix} \sigma_u^2 & \rho_{uv}\sigma_u \sigma_v \\ \rho_{uv}\sigma_u \sigma_v & \sigma_v^2 \end{bmatrix} \right)$

* This just says that both $u_i$ and $v_i$ come from a normal distribution
  * the variance of $u_i$ is $\sigma_u^2$
  * the variance of $v_i$ is $\sigma_v^2$
  * the correlation between $u_i$ and $v_i$ is $\rho_{uv}$
* *Note:* the population model includes the distribution of the random effects and noise

]

---

### Class activity

.large[
Work on the class activity on the course website. Submit your rendered HTML on Canvas at the end of class.
]




