---
title: Negative binomial regression
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Last time: class activity

```
            Estimate Std. Error t value Pr(>|t|)    
(Intercept)  2.837192   0.115099  24.650  < 2e-16 ***
male         0.456170   0.026095  17.481  < 2e-16 ***
age         -0.006806   0.001595  -4.267 2.07e-05 ***
education2   0.016463   0.029675   0.555    0.579    
education3   0.016427   0.037706   0.436    0.663    
education4  -0.015033   0.040477  -0.371    0.710    
diabetes    -0.025398   0.092606  -0.274    0.784    
BMI          0.005001   0.003304   1.513    0.130  
```

.large[
.question[
Is there a difference in the average number of cigarettes smoked between smokers with some college education and smokers with a college degree, after accounting for sex, age, diabetes, and BMI?
]
]

---

### Last time: class activity

.large[
$$Var(aU + bV) = a^2 Var(U) + b^2 Var(V) + 2ab Cov(U, V)$$
]

---

### Last time: class activity

```{r}
se <- sqrt(1.421765e-03 + 8.806037e-04 - 2*4.199471e-04) # 0.0382
t_stat <- (0.016427 - 0.016463)/se  # -0.001

2*pt(abs(t_stat), 2004, lower.tail=F)
```

---

### Comparing Poisson and quasi-Poisson

.large[
**Poisson:**

* Mean = $\lambda_i$
* Variance = $\lambda_i$

**quasi-Poisson:**

* Mean = $\lambda_i$
* Variance = $\phi \lambda_i$
* Variance is a linear function of the mean

.question[
What if we want variance to depend on the mean in a different way?
]
]

---

### Introducing the negative binomial

.large[
If $Y_i \sim NB(\theta, p)$, then $Y_i$ takes values $y = 0, 1, 2, 3, ...$ with probabilities

$$P(Y_i = y) = \dfrac{(y + \theta - 1)!}{y!(\theta - 1)!} (1 - p)^\theta p^y$$

* $\theta > 0$, $\ \ \ p \in [0, 1]$
* Mean = $\dfrac{p \theta}{1 - p} = \mu$
* Variance = $\dfrac{p \theta}{(1 - p)^2} = \mu + \dfrac{\mu^2}{\theta}$
* Variance is a *quadratic* function of the mean
]

---

### Mean and variance for a negative binomial variable

.large[
If $Y_i \sim NB(\theta, p)$, then

* Mean = $\dfrac{p \theta}{1 - p} = \mu$
* Variance = $\dfrac{p \theta}{(1 - p)^2} = \mu + \dfrac{\mu^2}{\theta}$

.question[
How is $\theta$ related to overdispersion?
]
]

---

### Negative binomial regression

.large[
$$Y_i \sim NB(\theta, \ p_i)$$

$$\log(\mu_i) = \beta_0 + \beta_1 X_i$$

* $\mu_i = \dfrac{p_i \theta}{1 - p_i}$
* Note that $\theta$ is the same for all $i$
* Note that just like in Poisson regression, we model the average count
  * Interpretation of $\beta$s is the same as in Poisson regression
]

---

### Comparing Poisson, quasi-Poisson, negative binomial

.large[
**Poisson:**

* Mean = $\lambda_i$
* Variance = $\lambda_i$

**quasi-Poisson:**

* Mean = $\lambda_i$
* Variance = $\phi \lambda_i$

**negative binomial:**

* Mean = $\mu_i$
* Variance = $\mu_i + \dfrac{\mu_i^2}{\theta}$
]

---

### In R

.large[
```{r, include=F}
library(foreign)
library(MASS)
library(tidyverse)

heart_data <- read.csv("~/Documents/Teaching/applied_glms/sta214-f25.github.io/class_activities/framingham.csv")

heart_data <- heart_data |>
  drop_na(male, age, education, diabetes, cigsPerDay, BMI) |>
  mutate(education = as.factor(education))

smokers <- heart_data |>
  filter(currentSmoker == 1)

m3 <- glm.nb(cigsPerDay ~ male + age + education + 
            diabetes + BMI, data = smokers)
```


```{r eval=F}
library(MASS)
m3 <- glm.nb(cigsPerDay ~ male + age + education + 
            diabetes + BMI, data = smokers)
```

```
             Estimate Std. Error z value Pr(>|z|)    
(Intercept)  2.877771   0.123477  23.306  < 2e-16 ***
male         0.459148   0.027641  16.611  < 2e-16 ***
age         -0.007010   0.001731  -4.050 5.12e-05 ***
education2   0.024518   0.032534   0.754    0.451    
education3   0.009252   0.040802   0.227    0.821    
education4  -0.027732   0.044825  -0.619    0.536    
diabetes    -0.010124   0.099126  -0.102    0.919    
BMI          0.003693   0.003573   1.033    0.301    

(Dispersion parameter for Negative Binomial(3.2981) family taken to be 1)
```

$\widehat{\theta} = 3.30$
]

---

### In R

.large[
```
             Estimate Std. Error z value Pr(>|z|)    
(Intercept)  2.877771   0.123477  23.306  < 2e-16 ***
male         0.459148   0.027641  16.611  < 2e-16 ***
age         -0.007010   0.001731  -4.050 5.12e-05 ***
education2   0.024518   0.032534   0.754    0.451    
education3   0.009252   0.040802   0.227    0.821    
education4  -0.027732   0.044825  -0.619    0.536    
diabetes    -0.010124   0.099126  -0.102    0.919    
BMI          0.003693   0.003573   1.033    0.301    
```

.question[
How do I interpret the estimated coefficient -0.007?
]
]

---

### Hypothesis testing

```
             Estimate Std. Error z value Pr(>|z|)    
(Intercept)  2.877771   0.123477  23.306  < 2e-16 ***
male         0.459148   0.027641  16.611  < 2e-16 ***
age         -0.007010   0.001731  -4.050 5.12e-05 ***
education2   0.024518   0.032534   0.754    0.451    
education3   0.009252   0.040802   0.227    0.821    
education4  -0.027732   0.044825  -0.619    0.536    
diabetes    -0.010124   0.099126  -0.102    0.919    
BMI          0.003693   0.003573   1.033    0.301  
```

.large[
**Research question:** Is there a relationship between education level and the number of cigarettes smoked per day, after accounting for sex, age, diabetes, and BMI?
]

---

### Likelihood ratio test

.large[
```{r}
m3 <- glm.nb(cigsPerDay ~ male + age + education + 
               diabetes + BMI, data = smokers)
m4 <- glm.nb(cigsPerDay ~ male + age +
               diabetes + BMI, data = smokers)
m3$twologlik - m4$twologlik
pchisq(1.423, df=3, lower.tail=F)
```
]


---

### quasi-Poisson vs. negative binomial

.large[

.pull-left[
**quasi-Poisson:**

* linear relationship between mean and variance
* easy to interpret $\widehat{\phi}$
* same as Poisson regression when $\phi = 1$
* simple adjustment to estimated standard errors
* estimated coefficients same as in Poisson regression
]

.pull-right[
**negative binomial:**

* quadratic relationship between mean and variance
* we get to use a likelihood, rather than a quasi-likelihood
* Same as Poisson regression when $\theta$ is very large and $p$ is very small
]
]

---

### Class activity

.large[
Work on the class activity. Submit the rendered HTML on Canvas at the end of class.
]

---

### Class activity

.large[
```{r, eval=F}
m_art <- glm.nb(art ~ ., data = articles)
```

```
             Estimate Std. Error z value Pr(>|z|)    
(Intercept)  0.256144   0.137348   1.865 0.062191 .  
femWomen    -0.216418   0.072636  -2.979 0.002887 ** 
marMarried   0.150489   0.082097   1.833 0.066791 .  
kid5        -0.176415   0.052813  -3.340 0.000837 ***
phd          0.015271   0.035873   0.426 0.670326    
ment         0.029082   0.003214   9.048  < 2e-16 ***
```
]

---

### Class activity

.large[
```{r, include=F}
library(MASS)
library(foreign)

articles <- read.dta("http://www.stata-press.com/data/lf2/couart2.dta")
```

```{r}
m_art <- glm.nb(art ~ ., data = articles)
m_art_reduced <- glm.nb(art ~ fem + mar + 
                          kid5 + ment,
                        data = articles)
m_art$twologlik - m_art_reduced$twologlik
pchisq(0.180, df=1, lower.tail=F)
```


]




