---
title: Logistic regression assumptions and diagnostics
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Motivating example: Dengue data

.large[
**Data:** Data on 5720 Vietnamese children, admitted to the hospital with possible dengue fever. Variables include:

* *Sex*: patient's sex (female or male)
* *Age*: patient's age (in years)
* *WBC*: white blood cell count
* *PLT*: platelet count
* other diagnostic variables...
* *Dengue*: whether the patient has dengue (0 = no, 1 = yes)
]

---

### Previously: Logistic regression model

.large[
$$Y_i = \text{dengue status (0 = negative, 1 = positive)}$$

$$Y_i \sim Bernoulli(\pi_i)$$

$$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 WBC_i$$

.question[
What assumptions does this logistic regression model make? How should we assess these assumptions? Discuss with your neighbor for 2--3 minutes, then we will discuss as a group.
]
]

---

### Residual plots for linear regression

.large[
$$PLT_i = \beta_0 + \beta_1 WBC_i + \varepsilon_i$$
]

```{r echo=F, message=F, fig.width=10, fig.height = 5, fig.align='center'}
library(tidyverse)
library(gridExtra)
dengue <- read.csv("https://sta279-s22.github.io/labs/dengue.csv")

m1 <- lm(PLT ~ WBC, data = dengue)
p1 <- dengue |>
  ggplot(aes(x = WBC, y = PLT)) +
  geom_point(size=1.5) +
  geom_smooth(se=F, method="lm", formula = y ~ x, color="blue", lwd=1.2) +
  theme_bw() +
  theme(text = element_text(size = 20))

p2 <- data.frame(WBC = dengue$WBC,
           Residual = dengue$PLT - m1$fitted.values) %>%
  ggplot(aes(x = WBC, y = Residual)) +
  geom_point(size=1.5) +
  geom_abline(slope = 0, intercept = 0, color="blue", lwd=1.2) +
  theme_bw() +
  theme(text = element_text(size = 20))

grid.arrange(p1, p2, ncol=2)
```




---

### Don't use raw residuals for logistic regression

.large[
Fitted model: $\log \left(\dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i}\right) = 1.737 - 0.361 \ WBC_i$
]

.large[
Raw residuals $Y_i - \widehat{\pi}_i$:
]

```{r echo=F, message=F, fig.width=7, fig.height = 5, fig.align='center'}


m1 <- glm(Dengue ~ WBC, data = dengue, family = binomial)
data.frame(WBC = dengue$WBC,
           Residual = dengue$Dengue - m1$fitted.values) %>%
  ggplot(aes(x = WBC, y = Residual)) +
  geom_point(size=1.5) +
  geom_abline(slope = 0, intercept = 0, color="blue", lwd=1.2) +
  theme_bw() +
  theme(text = element_text(size = 20))
```

---

### Assessing shape with empirical logit plots

.large[
**Example:** Putting data. Interested in the relationship between the length of a putt, and whether it was made:

$$Y_i \sim Bernoulli(\pi_i)$$

$$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 \ Length_i$$

| Length | 3 | 4 | 5 | 6 | 7 |
| --- | --- | --- | --- | --- | --- |
| Number of successes | 84 | 88 | 61 | 61 | 44 |
| Number of failures | 17 | 31 | 47 | 64 | 90 |
| Total | 101 | 119 | 108 | 125 | 134 |
]

---

### Empirical logits

.large[
**Step 1:** estimate the probability of success for each length of putt

| Length | 3 | 4 | 5 | 6 | 7 |
| --- | --- | --- | --- | --- | --- |
| Number of successes | 84 | 88 | 61 | 61 | 44 |
| Number of failures | 17 | 31 | 47 | 64 | 90 |
| Total | 101 | 119 | 108 | 125 | 134 |
| Probability of success $\widehat{\pi}$ | 0.832 | 0.739 | 0.565 | 0.488 | 0.328 |
]

---

### Empirical logits

.large[
**Step 2:** convert empirical probabilities to empirical log odds

| Length | 3 | 4 | 5 | 6 | 7 |
| --- | --- | --- | --- | --- | --- |
| Number of successes | 84 | 88 | 61 | 61 | 44 |
| Number of failures | 17 | 31 | 47 | 64 | 90 |
| Total | 101 | 119 | 108 | 125 | 134 |
| Probability of success $\widehat{\pi}$ | 0.832 | 0.739 | 0.565 | 0.488 | 0.328 |
| Odds $\dfrac{\widehat{\pi}}{1 - \widehat{\pi}}$ | 4.941 | 2.839 | 1.298 | 0.953 | 0.489 |
| Log-odds $\log \left( \dfrac{\widehat{\pi}}{1 - \widehat{\pi}} \right)$ | 1.60 | 1.04 | 0.26 | -0.05 | -0.72 |
]

---

### Empirical logits

.large[
**Step 3:** plot empirical log-odds against predictor, and add a least-squares line

```{r echo=F, message=F, fig.width=7, fig.height=4.5, fig.align='center'}
data.frame(length = c(3, 4, 5, 6, 7),
  log_odds = c(1.6, 1.04, 0.26, -0.05, -0.72)) %>% 
  ggplot(aes(x = length, y = log_odds)) + 
  geom_point(size=3) + 
  geom_smooth(se=F, method="lm") +
  labs(x = "Length (feet)",
       y = "Empirical log-odds") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.question[
Does it seem reasonable that the log-odds are a linear function of length?
]
]

---

### Back to the dengue data...

.large[

| WBC | 0.90 | 1.15 | 1.23 | 1.25 | 1.54 | 1.58 | ... |
| --- | --- | --- | --- | --- | --- | --- | --- |
| Dengue = 0 | 0 | 0 | 0 | 0 | 0 | 0 | ... |
| Dengue = 1 | 1 | 2 | 1 | 1 | 3 | 1 | ... |

.question[
What problem do I run into?
]
]

---

### Binned empirical logit plots

```{r echo=F, message=F, fig.align='center', fig.width=7, fig.height=5}
logodds_plot <- function(data, reg_formula, group,
                         num_bins = 5){
  
  var_names <- all.vars(reg_formula)

  if(length(var_names) != 2){
    stop("Only one variable can appear on the x-axis")
  }

  if(!is.numeric(data[[var_names[2]]])){
    stop("The explanatory variable must be numeric")
  }
  
  # to do: error message if outcome is not binary
  
  data <- data |>
    select(all_of(var_names), {{group}}) |>
    drop_na()

  dat <- model.frame(reg_formula, data)
  

  xname <- colnames(dat)[2]
  yname <- colnames(dat)[1]
  colnames(dat) <- c("y", "x")

  if(any(is.infinite(dat$x))){
    stop(paste("There are non-finite values in", xname))
  }

  if(!missing(group)){
    dat <- dat |>
      mutate({{group}} := as.factor(data |> pull({{group}})))
  }

  dat |>
    arrange({{group}}, x) |>
    group_by({{group}}) |>
    mutate(bin = rep(1:num_bins,
                     each=ceiling(n()/num_bins))[1:n()]) |>
    group_by(bin, {{group}}) |>
    summarize(x = mean(x),
                prop = (sum(y) + 0.5)/(n() + 1),
                num_obs = n(),
                .groups = "drop") |>
    mutate(logodds = log(prop/(1 - prop))) |>
    ggplot(aes(x = x,
               y = logodds,
               color = {{group}})) +
    geom_point() +
    geom_smooth(se=F, method="lm", formula = y ~ x) +
    theme_bw() +
    labs(x = xname,
         y = "Empirical log odds")
}

dengue |>
  logodds_plot(Dengue ~ WBC, num_bins = 15) +
  theme(text = element_text(size=20))
```

---

### Binned empirical logit plots

```{r echo=F, message=F, fig.align='center', fig.width=7, fig.height=5}
dengue |>
  logodds_plot(Dengue ~ WBC, num_bins = 15) +
  theme(text = element_text(size=20))
```

.large[
.question[
Does it seem reasonable that the log-odds are a linear function of WBC?
]
]

---

### Class activity, Part I

.large[
[https://sta214-f25.github.io/class_activities/ca_7.html](https://sta214-f25.github.io/class_activities/ca_7.html)
]

---

### Class activity

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=5}
titanic <- read.csv("https://sta214-s23.github.io/homework/Titanic.csv")

titanic |>
  logodds_plot(Survived ~ Fare, num_bins = 20) +
  theme(text = element_text(size=20))
```

---

### Class activity

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=5}
titanic |>
  logodds_plot(Survived ~ log(Fare + 1),
               num_bins = 20) +
  theme(text = element_text(size=20))
```

---

### Why residuals in linear regression are nice

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=7}
x <- runif(100, -1, 1)
y <- x + rnorm(100, sd=0.2)

p1 <- data.frame(x, y) |>
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  labs(title = "Dataset 1")

m1 <- lm(y ~ x)
p2 <- data.frame(x = x,
           Residual = y - m1$fitted.values) %>%
  ggplot(aes(x = x, y = Residual)) +
  geom_point(size=1.5) +
  geom_abline(slope = 0, intercept = 0, color="blue", lwd=1.2) +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  labs(title = "Residuals 1")


y <- x^2 + rnorm(100, sd=0.2)

p3 <- data.frame(x, y) |>
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  labs(title = "Dataset 2")

m1 <- lm(y ~ x)
p4 <- data.frame(x = x,
           Residual = y - m1$fitted.values) %>%
  ggplot(aes(x = x, y = Residual)) +
  geom_point(size=1.5) +
  geom_abline(slope = 0, intercept = 0, color="blue", lwd=1.2) +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  labs(title = "Residuals 2")


grid.arrange(p1, p3, p2, p4, ncol = 2)
```


---

### Quantile residuals for logistic regression

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=10, fig.height=4}
library(statmod)
set.seed(12)

m1 <- glm(Survived ~ Fare, data = titanic, family = binomial)
p1 <- data.frame(Fare = titanic$Fare, residuals = qresid(m1)) %>%
  ggplot(aes(x = Fare, y = residuals)) +
  geom_point(alpha = 0.3) +
  geom_smooth(lwd = 1.2, span=0.75) +
  theme_bw() +
  labs(title = "Survived ~ Fare",
       y = "Quantile residuals") +
  scale_x_continuous(limits = c(0, 200)) +
  theme(text = element_text(size = 20))


m1 <- glm(Survived ~ log(Fare + 1), data = titanic, family = binomial)
p2 <- data.frame(Fare = log(titanic$Fare + 1), residuals = qresid(m1)) %>%
  ggplot(aes(x = Fare, y = residuals)) +
  geom_point(alpha = 0.3) +
  geom_smooth(lwd = 1.2, span=0.75) +
  theme_bw() +
  labs(title = "Survived ~ log(Fare+1)",
       x = "log(Fare + 1)",
       y = "Quantile residuals") +
  theme(text = element_text(size = 20))

grid.arrange(p1, p2, ncol=2)
```


---

### Class activity, Part II

.large[
[https://sta214-f25.github.io/class_activities/ca_7.html](https://sta214-f25.github.io/class_activities/ca_7.html)
]
