---
title: Zero inflated models
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Data: College drinking

.large[
Survey data from 77 college students on a dry campus (i.e., alcohol is prohibited) in the US. Survey asks students "How many alcoholic drinks did you consume last weekend?"

* `drinks`: the number of drinks the student reports consuming
* `sex`: an indicator for whether the student identifies as male
* `OffCampus`: an indicator for whether the student lives off campus
* `FirstYear`: an indicator for whether the student is a first-year student

.question[
Our goal: model the number of drinks students report consuming.
]
]

---

### Class activity: The fitted model

.large[
$$P(Y_i = y) = \begin{cases} e^{-\lambda_i}(1 - \alpha_i) + \alpha_i & y = 0 \\ \dfrac{e^{-\lambda_i} \lambda_i^y}{y!}(1 - \alpha_i) & y > 0 \end{cases}$$

$\log \left( \dfrac{\widehat{\alpha}_i}{1 - \widehat{\alpha}_i} \right) = -0.60 + 1.14 FirstYear_i$

$\log(\widehat{\lambda}_i) = 0.75 + 0.42 \ OffCampus_i + 1.02 \ Male_i$

.question[

What is the estimated probability that a first year student never drinks?

]
]

---

### The fitted model

.large[
$$P(Y_i = y) = \begin{cases} e^{-\lambda_i}(1 - \alpha_i) + \alpha_i & y = 0 \\ \dfrac{e^{-\lambda_i} \lambda_i^y}{y!}(1 - \alpha_i) & y > 0 \end{cases}$$

$\log \left( \dfrac{\widehat{\alpha}_i}{1 - \widehat{\alpha}_i} \right) = -0.60 + 1.14 FirstYear_i$

$\log(\widehat{\lambda}_i) = 0.75 + 0.42 \ OffCampus_i + 1.02 \ Male_i$

.question[

What is the estimated average number of drinks for a male student who lives off campus and sometimes drinks?

]
]

---

### The fitted model

.large[
$$P(Y_i = y) = \begin{cases} e^{-\lambda_i}(1 - \alpha_i) + \alpha_i & y = 0 \\ \dfrac{e^{-\lambda_i} \lambda_i^y}{y!}(1 - \alpha_i) & y > 0 \end{cases}$$

$\log \left( \dfrac{\widehat{\alpha}_i}{1 - \widehat{\alpha}_i} \right) = -0.60 + 1.14 FirstYear_i$

$\log(\widehat{\lambda}_i) = 0.75 + 0.42 \ OffCampus_i + 1.02 \ Male_i$

.question[

What is the estimated probability that a male first year student who lives off campus had at least one drink last weekend?

]
]

---

```{r include=F}
library(tidyverse)
library(knitr)
library(pscl)
```


### Data: Framingham heart study

.large[
Data collected on residents of Framingham, MA over a long period of time, to study variables related to heart health. We will work with a subset of the data, containing

* `cigsPerDay`: The number of cigarettes smoked per day during the study period.
* `education`: 1 = High School, 2 = Some College, 3 = College Degree, 4 = Advanced Degree.
* `male`: 1 = Male, 0 = Female.
* `age`: The age of the individual in years.
* `diabetes`: 1 if the individual has diabetes, 0 otherwise.

.question[
Why might we see zero inflation in the number of cigarettes smoked per day?
]

]

---

### EDA: number of cigarettes smoked

```{r, echo=F, warning=F, message=F, fig.align='center', fig.height=5, fig.width=7}
heart_data <- read.csv("~/Documents/Teaching/applied_glms/sta214-f25.github.io/class_activities/framingham.csv")

heart_data <- heart_data %>%
  drop_na(male, age, education, diabetes, cigsPerDay) %>%
  mutate(education = as.factor(education))

heart_data %>%
  ggplot(aes(x = cigsPerDay)) +
  geom_histogram(bins = 20) +
  theme_bw() +
  labs(x = "Number of cigarettes smoked per day") +
  theme(text = element_text(size = 20))
```

---

### Class activity

.large[
Work on the class activity (handout).
]

---

### Class activity

$$P(Y_i = y) = \begin{cases} e^{-\lambda_i}(1 - \alpha_i) + \alpha_i & y = 0 \\ \dfrac{e^{-\lambda_i} \lambda_i^y}{y!}(1 - \alpha_i) & y > 0 \end{cases}$$

$$\log \left( \dfrac{\widehat{\alpha}_i}{1 - \widehat{\alpha}_i} \right) = -2.51 + 0.051 Age_i$$

$$\log(\widehat{\lambda}_i) = 3.21 -0.044 EducationSome_i -0.082 EducationCollege_i - \\ 0.0062 EducationAdv_i - 0.024 Diabetes_i - 0.0056 Age_i$$

.large[
.question[
How do we interpret the coefficient -0.024 in the fitted model?
]
]


---

### Class activity


$$\log \left( \dfrac{\widehat{\alpha}_i}{1 - \widehat{\alpha}_i} \right) = -2.51 + 0.051 Age_i$$

$$\log(\widehat{\lambda}_i) = 3.21 -0.044 EducationSome_i -0.082 EducationCollege_i - \\ 0.0062 EducationAdv_i - 0.024 Diabetes_i - 0.0056 Age_i$$

.large[
.question[
What is the expected number of cigarettes smoked per day, for a 50 year old smoker with diabetes and some college education?
]
]

---

### Making predictions


$$\log \left( \dfrac{\widehat{\alpha}_i}{1 - \widehat{\alpha}_i} \right) = -2.51 + 0.051 Age_i$$

$$\log(\widehat{\lambda}_i) = 3.21 -0.044 EducationSome_i -0.082 EducationCollege_i - \\ 0.0062 EducationAdv_i - 0.024 Diabetes_i - 0.0056 Age_i$$

.large[
.question[
What if I want the expected number of cigarettes smoked per day for a 50 year old with diabetes and some college education, but *without* knowing whether they are a smoker?
]
]

---

### A new question

$$P(Y_i = y) = \begin{cases} e^{-\lambda_i}(1 - \alpha_i) + \alpha_i & y = 0 \\ \dfrac{e^{-\lambda_i} \lambda_i^y}{y!}(1 - \alpha_i) & y > 0 \end{cases}$$

$$\log \left( \dfrac{\alpha_i}{1 - \alpha_i} \right) = \gamma_0 + \gamma_1 Age_i$$

$$\log(\lambda_i) = \beta_0 + \beta_1 EducationSome_i + \beta_2 EducationCollege_i + \\ \beta_3 EducationAdv_i + \beta_4 Diabetes_i + \beta_5 Age_i$$
.large[
.question[
New research question: for smokers, does the number of cigarettes smoked per day depend on age?

How would we answer this research question?
]
]

---

### Fitting the model in R

.large[
```{r, eval=F}
m2 <- zeroinfl(cigsPerDay ~ education + 
                 diabetes + age | age, 
               data = heart_data)
summary(m2)
```
]

```
Count model coefficients (poisson with log link):
              Estimate Std. Error z value Pr(>|z|)    
(Intercept)  3.2063376  0.0342290  93.673  < 2e-16 ***
education2  -0.0441182  0.0124809  -3.535 0.000408 ***
education3  -0.0820372  0.0158604  -5.172 2.31e-07 ***
education4  -0.0062453  0.0171640  -0.364 0.715964    
diabetes    -0.0241459  0.0386337  -0.625 0.531974    
age         -0.0056182  0.0006738  -8.338  < 2e-16 ***

Zero-inflation model coefficients (binomial with logit link):
             Estimate Std. Error z value Pr(>|z|)    
(Intercept) -2.506681   0.191483  -13.09   <2e-16 ***
age          0.051329   0.003821   13.43   <2e-16 ***
```

---

### z-test

.large[
```{r, eval=F}
m2 <- zeroinfl(cigsPerDay ~ education + 
                 diabetes + age | age, 
               data = heart_data)
summary(m2)
```
]

```
Count model coefficients (poisson with log link):
              Estimate Std. Error z value Pr(>|z|)    
(Intercept)  3.2063376  0.0342290  93.673  < 2e-16 ***
education2  -0.0441182  0.0124809  -3.535 0.000408 ***
education3  -0.0820372  0.0158604  -5.172 2.31e-07 ***
education4  -0.0062453  0.0171640  -0.364 0.715964    
diabetes    -0.0241459  0.0386337  -0.625 0.531974    
age         -0.0056182  0.0006738  -8.338  < 2e-16 ***
```
---

### Likelihood ratio test

```{r, include=F}
m2 <- zeroinfl(cigsPerDay ~ education + 
                 diabetes + age | age, 
               data = heart_data)
summary(m2)
```

---

### Likelihood ratio test

.large[

```{r}
m2 <- zeroinfl(cigsPerDay ~ education + 
                 diabetes + age | age, 
               data = heart_data)
m2$loglik
```

```{r}
m1 <- zeroinfl(cigsPerDay ~ education + 
                 diabetes | age, 
               data = heart_data)
m1$loglik
```
]

---

### Likelihood ratio test

.large[

```{r}
m1 <- zeroinfl(cigsPerDay ~ education + 
                 diabetes | age, 
               data = heart_data)

m2 <- zeroinfl(cigsPerDay ~ education + 
                 diabetes + age | age, 
               data = heart_data)

2*(m2$loglik - m1$loglik)
pchisq(69.976, df=1, lower.tail=F)
```
]

---

### Class activity

.large[
Work on the class activity. At the end of class, render your work as an HTML and submit on Canvas.
]


