---
title: Quasi-Poisson models 
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---


### Last time: modeling article publication

```{r include=F}
library(knitr)
library(tidyverse)
library(MASS)
library(foreign)
articles <- read.dta("http://www.stata-press.com/data/lf2/couart2.dta")


hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })
```

.large[
We are interested in analyzing the number of articles published by biochemistry PhD students. The data contains the following variables:

* `art`: articles published in last three years of Ph.D.
* `fem`: gender (recorded as male or female)
* `mar`: marital status (recorded as married or single)
* `kid5`: number of children under age six
* `phd`: prestige of Ph.D. program
* `ment`: articles published by their mentor in last three years

$$Articles_i \sim Poisson(\lambda_i)$$
$$\log(\lambda_i) = \beta_0 + \beta_1 Female_i + \beta_2 Married_i + \beta_3 Kids_i + \\  \beta_4 Prestige_i + \beta_5 Mentor_i$$
]

---

### Recap: confidence interval

.large[
```
             Estimate Std. Error z value Pr(>|z|)    
(Intercept)  0.304617   0.102981   2.958   0.0031 ** 
femWomen    -0.224594   0.054613  -4.112 3.92e-05 ***
marMarried   0.155243   0.061374   2.529   0.0114 *  
kid5        -0.184883   0.040127  -4.607 4.08e-06 ***
phd          0.012823   0.026397   0.486   0.6271    
ment         0.025543   0.002006  12.733  < 2e-16 ***
```

Confidence interval for $\beta_4$:
]

---

### Goodness of fit test

.large[
```
    Null deviance: 1817.4  on 914  degrees of freedom
Residual deviance: 1634.4  on 909  degrees of freedom
```
]

---

### Offest?

.large[
* `art`: articles published in last three years of Ph.D.
* `fem`: gender (recorded as male or female)
* `mar`: marital status (recorded as married or single)
* `kid5`: number of children under age six
* `phd`: prestige of Ph.D. program
* `ment`: articles published by their mentor in last three years

$$Articles_i \sim Poisson(\lambda_i)$$
$$\log(\lambda_i) = \beta_0 + \beta_1 Female_i + \beta_2 Married_i + \beta_3 Kids_i + \\  \beta_4 Prestige_i + \beta_5 Mentor_i$$

.question[
Do you think this model needs an offset?
]
]

---

### Overdispersion

.large[
**Overdispersion** occurs when the response $Y$ has higher variance than we would expect if $Y$ followed a Poisson distribution.

Formally, let

$$\phi = \dfrac{\text{Variance}}{\text{Mean}}$$
.question[
What should $\phi$ be if there is no overdispersion?
]
]

---

### Estimating overdispersion

.large[
The *Pearson residual* for observation $i$ is defined as

$$e_{(P)i} = \dfrac{Y_i - \widehat{\lambda}_i}{\sqrt{\widehat{\lambda}_i}}$$
]

</br>

.large[
$$\widehat{\phi} = \dfrac{\sum \limits_{i=1}^n e_{(P)i}^2}{n - p}$$

* $p =$ number of parameters in model
]

---

### Example: Estimating overdispersion

.large[
```{r}
# fit the model
m1 <- glm(art ~ ., data = articles, 
          family = poisson)

# get Pearson residuals
pearson_resids <- resid(m1, "pearson")

# estimate dispersion parameter
phihat <- sum(pearson_resids^2)/(915 - 6)
phihat
```

]

---


### Handling overdispersion

.large[
Overdispersion is a problem because our standard errors (for confidence intervals and hypothesis tests) are too low.

.question[
If we think there is overdispersion, what should we do?
]
]

---

### Adjusting the standard error

.large[
* In our data, $\widehat{\phi} = 1.83$
* This means our variance is 1.83 times bigger than it should be
* So our standard error is $\sqrt{1.83} = 1.35$ times bigger than it should be
]

---

### Adjusting the standard error in R

.large[
```{r, message=F, warning=F}
m2 <- glm(art ~ ., data = articles, 
          family = quasipoisson)
```
]

```{r echo=F, output.lines = 10:20}
summary(m2)
```

.large[
* Allowing $\phi$ to be different from 1 means we are using a *quasi-likelihood* (in this case, a *quasi-Poisson*)
]

---

### t-distribution

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=10, fig.height=6}
x <- seq(-4, 4, 0.01)
y <- c(dt(x, 1), dt(x, 2), dt(x, 5), dnorm(x))
df <- rep(c("df = 1", "df=2", "df=5", "N(0, 1)"), each = length(x))
data.frame(x = rep(x, 4), y, df) %>%
  ggplot(aes(x = x, y = y, color = df)) +
  geom_line(lwd = 1) +
  labs(y = "Density", color="") +
  theme_bw() +
  theme(text = element_text(size = 20))
```


---

### Calculating a confidence interval

```{r echo=F, output.lines = 11:16}
summary(m2)
```

.large[
New confidence interval for $\beta_4$:

$0.0128 \pm t_{n-p}^* \cdot 0.0357$

```{r}
qt(0.025, df=909, lower.tail=F)
```

$0.0128 \pm 1.96 \cdot 0.0357 = (-0.0572, \ 0.0828)$
]

---

### Adjusting the standard error in R

.large[
**Poisson:**

```{r echo=F, output.lines = 10:14}
m1 <- glm(art ~ ., data = articles, 
          family = poisson)
summary(m1)
```

**Quasi-Poisson:**

```{r echo=F, output.lines = 10:14}
summary(m2)
```
]

---

### Class activity

.large[
Work on the class activity. Submit your rendered HTML on Canvas at the end of class.
]



