---
title: Deviance, AIC, and model building
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### SBA data

.large[
Data on loan defaults for US Small Business Administration (SBA) loans:

* `Default`: whether the business defaulted on the loan (1 = yes, 0 = no)
* `UrbanRural`: 1 if business is in urban area, 2 if business is in rural area, 0 if unknown
* `NewExist`: 1 if business already existed, 2 if business is new, 0 if unknown
* `Term`: Length of the loan term (months)
* `DisbursementGross`: The amount of money disbursed (loaned), in dollars
* Many other variables...
]

---

### Last time: Class activity

```{r, include=F}
sba <- read.csv("https://sta214-s23.github.io/class_activities/sba_small.csv")
```

.large[
```{r}
m1 <- glm(Default ~ log(DisbursementGross) + Term + 
            as.factor(UrbanRural), 
          family = binomial, data = sba)

table(Prediction = m1$fitted.values > 0.5, 
      Truth = sba$Default)
```

* Accuracy = 
* Sensitivity = 
* Specificity = 
* PPV = 
]

---

### Last time: Class activity

.large[
```{r, echo=F}
table(Prediction = m1$fitted.values > 0.5, 
      Truth = sba$Default)
```

.question[
Is an accuracy of around 80% good?
]
]

---

### Last time: Class activity

.large[
Changing thresholds:

```{r}
table(Prediction = m1$fitted.values > 0.3, 
      Truth = sba$Default)
```

```{r}
table(Prediction = m1$fitted.values > 0.7, 
      Truth = sba$Default)
```

]

---

### A new research question

.large[
* `Default`: whether the business defaulted on the loan (1 = yes, 0 = no)
* `UrbanRural`: 1 if business is in urban area, 2 if business is in rural area, 0 if unknown
* `NewExist`: 1 if business already existed, 2 if business is new, 0 if unknown
* `Term`: Length of the loan term (months)
* `DisbursementGross`: The amount of money disbursed (loaned), in dollars
* Many other variables...

.question[
**Research question:** Which combination of variables "best" models loan default?
]
]

---

### A new research question

.large[
.question[
**Research question:** Which combination of variables "best" models loan default?
]

We need:

* A metric to compare different models
* A way to efficiently search through many different models
]

---

### Logistic regression

.large[
**Data:** Grad application data
  * `admit`: accepted to grad school? (0 = no, 1 = yes)
  * `gre`: GRE score
  * `gpa`: undergrad GPA
  * `rank`: prestige of undergrad institution
  
$Admit_i \sim Bernoulli(\pi_i) \hspace{0.5cm} \log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 GPA_i$

**Goal:** Summarize how well our model fits the data

.question[
How do we summarize model fit in *linear* regression?
]
]

---

### Summarizing linear regression model fit

```{r, include=F}
library(tidyverse)
grad_app <- read.csv("https://stats.idre.ucla.edu/stat/data/binary.csv")

library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })
```

.large[
Example model:
]

```{r, output.lines = c(9:12, 16:17)}
linear_mod <- lm(gre ~ gpa, data = grad_app)
summary(linear_mod)
```

---

### Summarizing linear regression model fit

.large[
Example model:
]

```{r, output.lines = c(9:12, 16:17)}
linear_mod <- lm(gre ~ gpa, data = grad_app)
summary(linear_mod)
```

.large[
.question[
Is $R^2_{adj}$ appropriate for logistic regression?
]
]

---

### Summarizing logistic regression

.large[
```{r, output.lines = c(6:9, 14:15)}
logistic_mod <- glm(admit ~ gpa, data = grad_app,
                    family = binomial)
summary(logistic_mod)
```
]

---

### Deviance

.large[
```{r, output.lines = c(6:9, 14:15)}
logistic_mod <- glm(admit ~ gpa, data = grad_app,
                    family = binomial)
summary(logistic_mod)
```
]

---

### Deviance

.large[
Example:
]

```{r, output.lines=8:10}
glm(admit ~ gre, family = binomial, data = grad_app)
```

```{r, output.lines=8:10}
glm(admit ~ gpa, family = binomial, data = grad_app)
```

.large[
.question[
Which predictor (GRE or GPA) gives a model with a better fit?
]
]

---

### Comparing models


```{r, output.lines=8:10}
glm(admit ~ gre, family = binomial, data = grad_app)
```

```{r, eval=F}
glm(admit ~ gre + gpa, family = binomial, data = grad_app)
```

.large[
.question[
How will deviance change when I add another variable to the model?
]
]

---

### Comparing models

```{r, output.lines=8:10}
glm(admit ~ gre, family = binomial, data = grad_app)
```

```{r, output.lines=8:10}
glm(admit ~ gre + gpa, family = binomial, data = grad_app)
```

.large[
Deviance will *never* increase when adding more explanatory variables to a model!
]

---

### Comparing models

.large[
.question[
In linear regression, what quantity did we use to compare models with different numbers of parameters?
]
]

---

### AIC

```{r, output.lines=8:10}
glm(admit ~ gre + gpa, family = binomial, data = grad_app)
```

.large[
Suppose our model has $p$ parameters (the number of $\beta$s, including the intercept). Then the AIC is

$$AIC = \text{deviance} + 2p$$
]

---

### AIC

.large[
$$AIC = \text{deviance} + 2p$$
]

**Model 1:** 

$$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 Term_i + \beta_2 \log(DisbursementGross_i)$$
.large[
`Residual Deviance: 3974 	AIC: 3980`
]

**Model 2:** 

$$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 Term_i + \beta_2 Urban_i + \beta_3 Rural_i + \beta_4 Employees_i$$
.large[
`Residual Deviance: 3827   AIC:`

.question[
Which model do we prefer, based on AIC?
]
]

---

### Back to the research question

.large[
.question[
**Research question:** Which combination of variables "best" models loan default?
]

We need:

* A metric to compare different models
  * Solution: AIC
* A way to efficiently search through many different models

.question[
How should we search many different models?
]
]

---

### Some model search algorithms

.large[
Best subset selection:

<br/>

<br/>

<br/>

Stepwise selection:
]

---

### Class activity

.large[
[https://sta214-f25.github.io/class_activities/ca_11.html](https://sta214-f25.github.io/class_activities/ca_11.html)
]

.large[
* Work independently or with a neighbor
* Submit HTML on Canvas when done
]






