---
title: Poisson Regression 
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Class activity: Poisson regression assumptions

.large[

[https://sta214-f25.github.io/class_activities/ca_21.html](https://sta214-f25.github.io/class_activities/ca_21.html)

Work on the class activity, then we will discuss as a class.
]

---

### Class activity: Poisson regression assumptions

```{r, echo=F, message=FALSE, warning=F, fig.align='center', fig.width=10, fig.height=7}
library(tidyverse)
library(statmod)
library(topmodels)
library(gridExtra)

set.seed(1)
r <- 0.5
n <- 1000
x <- rnorm(n, sd = 0.5)
y1 <- rpois(n, lambda = exp(x))
y2 <- rpois(n, lambda = exp(x^2))
y3 <- rnbinom(n, size=r, mu=exp(x))
y4 <- rnbinom(n, size=r, mu=exp(x^2))

m1 <- glm(y1 ~ x, family = poisson)
m2 <- glm(y2 ~ x, family = poisson)
m3 <- glm(y3 ~ x, family = poisson)
m4 <- glm(y4 ~ x, family = poisson)

p1 <- data.frame(x = x, resids = qresid(m1)) |>
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(x = "X", y = "Quantile residuals", 
       title = "Poisson data, shape assumption satisfied") +
  scale_y_continuous(limits=c(-3, 6))
p2 <- data.frame(x = x, resids = qresid(m2)) |>
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(x = "X", y = "Quantile residuals", 
       title = "Poisson data, shape assumption violated") +
  scale_y_continuous(limits=c(-3, 6))
p3 <- data.frame(x = x, resids = qresid(m3)) |>
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(x = "X", y = "Quantile residuals", 
       title = "Non-Poisson data, shape assumption satisfied") +
  scale_y_continuous(limits=c(-3, 6))
p4 <- data.frame(x = x, resids = qresid(m4)) |>
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(x = "X", y = "Quantile residuals", 
       title = "Non-Poisson data, shape assumption violated") +
  scale_y_continuous(limits=c(-3, 6))

grid.arrange(p1, p2, p3, p4, ncol=2)
```

---

### Using quantile residual plots

.large[
All assumptions are satisfied:

* Residuals are randomly scattered around 0, with no clear pattern
* Residuals have constant variance
* Residuals are mostly between -2 and 2

If you see a pattern in the residuals:

* Shape assumption may not be satisfied

If residuals have non-constant variance, and/or variance is too large:

* Distribution assumption may not be satisfied
]

---

### Rootograms

```{r, echo=F, fig.width=7, fig.height=4, fig.align="center"}
rootogram(m1, style = "standing")
```

---

### How are rootograms created?

```{r, echo=F, fig.width=6.5, fig.height=3.5, fig.align="center"}
rootogram(m1, style = "standing")
```

```{r}
sqrt(sum(y1 == 0))
```

```{r}
lambda_hat <- m1$fitted.values
sqrt(sum(lambda_hat^0 * exp(-lambda_hat)))
```

---

### Examining rootograms

```{r, echo=F, fig.width=6.5, fig.height=3.5, fig.align="center"}
rootogram(m1, style = "standing")
```

```{r, echo=F, fig.width=6.5, fig.height=3.5, fig.align="center"}
rootogram(m3, style = "standing")
```

---

### Examining rootograms

.large[
If all assumptions are satisfied (i.e., the model is a good fit to the data):

* The histogram bars will be close to the predictions (in red) from the fitted model

If the model is not a good fit to the data (wrong shape and/or distribution):

* The predictions will not match the observed frequencies
]

---

### Another dataset

.large[
A concerned parent asks us to investigate crime rates on college campuses. We have access to data on 81 different colleges and universities in the US, including the following variables:

* `type`: college (C) or university (U)
* `nv`: the number of crimes for that institution in the given year
* `enroll1000`: the number of enrolled students, in thousands
* `region`: region of the US C = Central, MW = Midwest, NE = Northeast, SE = Southeast, SW = Southwest, and W = West)
]

---

### Question

.large[
We want to know whether there are regional differences in the number of crimes on college campuses.

$$Crimes_i \sim Poisson(\lambda_i)$$

$$\log(\lambda_i) = \beta_0 + \beta_1 MW_i + \beta_2 NE_i + \beta_3 SE_i + \beta_4 SW_i + \beta_5 W_i$$
]

---

### Exploratory data analysis

.large[
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=10, fig.height=6}
library(tidyverse)

crimes <- read_csv("c_data.csv")

crimes |>
  ggplot(aes(x = nv)) +
  geom_histogram(bins = 10) +
  labs(x = "Number of crimes") +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  facet_wrap(~region)
```

.question[
Does it look reasonable to assume a Poisson distribution for the response?
]
]

---

### Exploratory data analysis

.large[
```{r}
crimes |>
  group_by(region) |>
  summarize(mean_crimes = mean(nv),
            var_crimes = var(nv))
```

.question[
Does the Poisson distribution still seem reasonable?
]
]

---

### Quantile residual plot

```{r}
m1 <- glm(nv ~ region, data = crimes, family = poisson)
```

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
crimes |>
  mutate(resids = qresid(m1)) |>
  ggplot(aes(x = region, y = resids)) +
  geom_point() +
  theme_bw() +
  labs(x = "Region", y = "Quantile residuals")  +
  theme(text = element_text(size = 20))
```

---

### Rootogram

```{r, fig.width=7, fig.height=5, fig.align='center'}
m1 <- glm(nv ~ region, data = crimes, family = poisson)
rootogram(m1, style="standing")
```

---

### Goodness of fit

.large[
Another way to assess whether our model is reasonable is with a *goodness of fit* test.

**Goodness of fit test:** If the model is a good fit for the data, then the residual deviance follows a $\chi^2$ distribution with the same degrees of freedom as the residual deviance

```{r echo=F, message=F, output.lines = 19:20}
library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

m1 <- glm(nv ~ region, data = crimes, family = poisson)
summary(m1)
```

Residual deviance = 621.24, df = 75

.question[
How likely is a residual deviance of 621.24 if our model is correct?
]
]

---

### Goodness of fit

.large[
Residual deviance = 621.24, df = 75
]

```{r, echo=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
x <- qchisq(ppoints(1000), 75)
y <- dchisq(x, 75)
data.frame(x, y) |>
  ggplot(aes(x = x, y = y)) +
  geom_line() +
  theme_bw() +
  labs(y = "density", title = "Chi-squared distribution with 75 df")
```

```{r}
pchisq(621.24, df=75, lower.tail=F)
```


---

### Goodness of fit

.large[
**Goodness of fit test:** If the model is a good fit for the data, then the residual deviance follows a $\chi^2$ distribution with the same degrees of freedom as the residual deviance

Residual deviance = 621.24, df = 75

```{r}
pchisq(621.24, df=75, lower.tail=F)
```

So our model might not be a very good fit to the data.

.question[
Why might our model not be a good fit?
]
]

---

### Potential issues with our model

.large[
* The Poisson distribution might not be a good choice
* There may be additional factors related to the number of crimes which we are not including in the model

.question[
Which other factors might be related to the number of crimes?
]
]



