---
title: Comparing models
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
grad_app <- read.csv("https://stats.idre.ucla.edu/stat/data/binary.csv")

library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })
```

### Hypothesis testing

.large[
$Y_i \sim Bernoulli(\pi_i) \hspace{0.5cm} \log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 \text{GRE}_i$

We want to know if there is actually a relationship between GRE score and grad school admission.

.question[
How can I express this as null and alternative hypotheses about one or more model parameters?
]
]

---

### Steps in hypothesis testing

---

### Test statistic

.large[
```{r, output.lines = c(4:8)}
m1 <- glm(admit ~ gre, family = binomial, 
          data = grad_app)
summary(m1)
```

.question[
Which part of this output should I use as my test statistic for the hypothesis test?
]
]

---

### p-value

.large[
```{r, output.lines = c(4:8)}
m1 <- glm(admit ~ gre, family = binomial, 
          data = grad_app)
summary(m1)
```

.question[
Which part of this output gives me the p-value for my hypothesis test?
]
]

---

### p-value

.large[
```{r, output.lines = c(4:8)}
m1 <- glm(admit ~ gre, family = binomial, 
          data = grad_app)
summary(m1)
```

.question[
What distribution are we using to calculate that p-value?
]
]

---

### p-value

.large[
```{r, output.lines = c(4:8)}
m1 <- glm(admit ~ gre, family = binomial, 
          data = grad_app)
summary(m1)
```

Our goal today: why is a N(0, 1) distribution appropriate for calculating the p-value here?
]

---

### Calculating p-values

.large[
.question[
What *is* a p-value?
]
]

---

### Null distribution

.large[
**Null distribution:** distribution of our test statistic *under* $H_0$

* We need to know how our test statistic behaves if the null hypothesis is true
* That is, how would the test statistic vary from sample to sample?
]

---

### Observed relationship

```{r, include=F}
logodds_plot <- function(data, reg_formula, group,
                         num_bins = 5){
  
  var_names <- all.vars(reg_formula)

  if(length(var_names) != 2){
    stop("Only one variable can appear on the x-axis")
  }

  if(!is.numeric(data[[var_names[2]]])){
    stop("The explanatory variable must be numeric")
  }
  
  if(n_distinct(data[[var_names[1]]]) != 2){
    stop("The response variable must be binary")
  }
  
  data <- data |>
    select(all_of(var_names), {{group}}) |>
    drop_na()

  dat <- model.frame(reg_formula, data)
  

  xname <- colnames(dat)[2]
  yname <- colnames(dat)[1]
  colnames(dat) <- c("y", "x")

  if(any(is.infinite(dat$x))){
    stop(paste("There are non-finite values in", xname))
  }

  if(!missing(group)){
    dat <- dat |>
      mutate({{group}} := as.factor(data |> pull({{group}})))
  }

  dat |>
    arrange({{group}}, x) |>
    group_by({{group}}) |>
    mutate(bin = rep(1:num_bins,
                     each=ceiling(n()/num_bins))[1:n()]) |>
    group_by(bin, {{group}}) |>
    summarize(x = mean(x),
                prop = (sum(y) + 0.5)/(n() + 1),
                num_obs = n(),
                .groups = "drop") |>
    mutate(logodds = log(prop/(1 - prop))) |>
    ggplot(aes(x = x,
               y = logodds,
               color = {{group}})) +
    geom_point() +
    geom_smooth(se=F, method="lm", formula = y ~ x) +
    theme_bw() +
    labs(x = xname,
         y = "Empirical log odds")
}
```

```{r, eval=F}
logodds_plot(grad_app, admit ~ gre, num_bins = 10)
```

```{r, echo=F, fig.width=6, fig.height=4, fig.align='center'}
logodds_plot(grad_app, admit ~ gre, num_bins = 10) +
  theme(text = element_text(size = 20))
```

.large[
.question[
What would this plot look like if $H_0$ were true?
]
]

---

### Making $H_0$ true

.large[
* Currently, there may be a relationship between GRE and Admit
* To remove this relationship: randomly shuffle the GRE values
]

---

### Making $H_0$ true

```{r, eval=F}
grad_app_perm <- grad_app |>
  mutate(gre = sample(grad_app$gre, n(), replace = F))

logodds_plot(grad_app_perm, admit ~ gre, num_bins = 10)
```

```{r, echo=F, fig.width=6, fig.height=4, fig.align='center'}
set.seed(9)

grad_app_perm <- grad_app |>
  mutate(gre = sample(grad_app$gre, n(), replace = F))

logodds_plot(grad_app_perm, admit ~ gre, num_bins = 10) +
  theme(text = element_text(size = 20))
```

---

### Making $H_0$ true

```{r, eval=F}
grad_app_perm <- grad_app |>
  mutate(gre = sample(grad_app$gre, n(), replace = F))

perm_mod <- glm(admit ~ gre, family = binomial, 
                  data = grad_app_perm)

summary(perm_mod)
```

```{r, echo=F, output.lines = c(4:8)}
set.seed(9)

grad_app_perm <- grad_app |>
  mutate(gre = sample(grad_app$gre, n(), replace = F))

perm_mod <- glm(admit ~ gre, family = binomial, 
                  data = grad_app_perm)

summary(perm_mod)
```

.large[
.question[
What is my test statistic for the permuted data (for which $H_0$ is *true*)?
]
]

---

### Making $H_0$ true

```{r, eval=F}
grad_app_perm <- grad_app |>
  mutate(gre = sample(grad_app$gre, n(), replace = F))

perm_mod <- glm(admit ~ gre, family = binomial, 
                  data = grad_app_perm)

summary(perm_mod)
```

```{r, echo=F, output.lines = c(4:8)}
set.seed(9)

grad_app_perm <- grad_app |>
  mutate(gre = sample(grad_app$gre, n(), replace = F))

perm_mod <- glm(admit ~ gre, family = binomial, 
                  data = grad_app_perm)

summary(perm_mod)
```

.large[
.question[
Does this give me the *distribution* of my test statistic under $H_0$?
]
]

---

### Repeating the process

.large[
```{r}
nreps <- 1000 # number of repetitions
z_stats <- rep(NA, nreps) # place to store results

head(z_stats)
```
]

---

### Repeating the process

.large[
```{r, eval=F}
nreps <- 1000 # number of repetitions
z_stats <- rep(NA, nreps) # place to store results

for(i in 1:nreps){
  # do something
}
```
]

---

### For loop: example

.large[
```{r}
for(i in 1:5){
  print(i)
}
```
]

---

### For loop: example

.large[
```{r}
ex_results <- rep(NA, 5)
for(i in 1:5){
  ex_results[i] <- i*2
}

ex_results
ex_results[1]
ex_results[2]
```
]

---

### Repeating the process

.large[
```{r, eval=F}
nreps <- 1000 # number of repetitions
z_stats <- rep(NA, nreps) # place to store results

for(i in 1:nreps){
  # do something
}
```

.question[
What needs to happen inside our for loop?
]
]

---

### Repeating the process

.large[
```{r, eval=F}
nreps <- 1000 # number of repetitions
z_stats <- rep(NA, nreps) # place to store results

for(i in 1:nreps){
  grad_app_perm <- grad_app |>
    mutate(gre = sample(grad_app$gre, n(), 
                        replace = F))
  
  perm_mod <- glm(admit ~ gre, family = binomial, 
                  data = grad_app_perm)
  z_stats[i] <- summary(perm_mod)$coefficients[2,3]
}
```
]


---

### Class activity

.large[
[https://sta214-f25.github.io/class_activities/ca_12.html](https://sta214-f25.github.io/class_activities/ca_12.html)
]

.large[
* Work independently or with a neighbor
* Submit HTML on Canvas when done
]




