---
title: Inference for logistic regression models
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Agenda

.large[
* For homework submissions:
    * Make sure to submit rendered HTML
    * If writing math in Quarto (e.g. regression models, maximum likelihood problems, etc.), make sure the equations display correctly
* Discuss feedback survey
* Recap activity from end of last class
* More inference!
]

---

### Last time: Activity

$$Admit_i \sim Bernoulli(\pi_i)$$

$$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 GPA_i + \beta_2 GRE_i + \beta_3 Rank2_i + \beta_4 Rank3_i + \beta_5 Rank4_i$$

.large[
.question[
We want to test whether there is a relationship between Rank and grad school acceptance, after accounting for GPA and GRE. What are my null and alternative hypotheses?
]
]

---

### Activity

$$Admit_i \sim Bernoulli(\pi_i)$$

$$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 GPA_i + \beta_2 GRE_i + \beta_3 Rank2_i + \beta_4 Rank3_i + \beta_5 Rank4_i$$

.large[
$H_0: \beta_3 = \beta_4 = \beta_5 = 0$

$H_A:$ at least one of $\beta_3, \beta_4, \beta_5 \neq 0$
]

.large[
.question[
What is the reduced model?
]
]

---

### Activity

```{r, echo=F}
grad_app <- read.csv("https://stats.idre.ucla.edu/stat/data/binary.csv")
glm(admit ~ gpa + gre + factor(rank), family = binomial, data = grad_app)
glm(admit ~ gpa + gre, family = binomial, data = grad_app)
```

---

### Activity

.large[
$H_0: \beta_3 = \beta_4 = \beta_5 = 0$

$H_A:$ at least one of $\beta_3, \beta_4, \beta_5 \neq 0$

Drop-in-deviance test: $G = 21.8$

.question[
What is the distribution of $G$ under the null hypothesis?
]
]

---

### Activity

.large[
$H_0: \beta_3 = \beta_4 = \beta_5 = 0$

$H_A:$ at least one of $\beta_3, \beta_4, \beta_5 \neq 0$

Drop-in-deviance test: $G = 21.8$

Under $H_0$, $G \sim \chi^2_3$

```{r}
pchisq(21.8, df=3, lower.tail=F)
```

]

---

### A new question

.large[
$$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 GRE_i$$

```
             Estimate Std. Error z value Pr(>|z|)    
(Intercept) -2.901344   0.606038  -4.787 1.69e-06
gre          0.003582   0.000986   3.633  0.00028
```

.question[
Here $\widehat{\beta}_1 = 0.003582$. Will $\widehat{\beta}_1$ be exactly equal to the true value $\beta_1$?
]
]

---

### A new question

.large[
$$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 GRE_i$$

```
             Estimate Std. Error z value Pr(>|z|)    
(Intercept) -2.901344   0.606038  -4.787 1.69e-06
gre          0.003582   0.000986   3.633  0.00028
```

.question[
Here $\widehat{\beta}_1 = 0.003582$. Will $\widehat{\beta}_1$ be exactly equal to the true value $\beta_1$?
]

No -- our estimate comes from one sample of data, and there is variability from sample to sample.

.question[
How can I account for this variability when reporting an estimate of $\beta_1$?
]
]

---

### Interval estimates

.large[
**Point estimate:** a single "best guess" value of a parameter. The estimated slope $\widehat{\beta}_1$ is a point estimate of $\beta_1$

**Interval estimate:** a "reasonable range" of possible values for a parameter. A *confidence interval* is an interval estimate

.question[
How would I construct a confidence interval for $\beta_1$?
]
]

---

### The critical value $z^*$

.large[
To construct a $(1 - \alpha)\cdot 100\%$ confidence interval, $z^*$ is the value such that $P(Z > z^*) = \alpha/2$
]

.pull-left[
```{r, echo=F, message=F, warning=F, fig.width=6, fig.height=4}
library(tidyverse)
x <- qnorm(ppoints(1000))
y <- dnorm(x)
data.frame(x, y) |>
  ggplot(aes(x = x, y = y)) +
  geom_line() +
  theme_classic() +
  geom_vline(xintercept = c(-1.96, 1.96), color="red") +
  labs(x = "", y = "", title = "Critical value for a 95% CI") +
  theme(text = element_text(size = 20))
```

]


.pull-right[
```{r, echo=F, fig.width=6, fig.height=4}
x <- qnorm(ppoints(1000))
y <- dnorm(x)
data.frame(x, y) |>
  ggplot(aes(x = x, y = y)) +
  geom_line() +
  theme_classic() +
  geom_vline(xintercept = c(-2.576, 2.576), color="red") +
  labs(x = "", y = "", title = "Critical value for a 99% CI") +
  theme(text = element_text(size = 20))
```
]

---

### Computing $z^*$


.large[
Example: for a 95% confidence interval, $z^*$ = 1.96

```{r}
qnorm(0.025, lower.tail=FALSE)
```

Example: for a 99% confidence interval, $z^*$ = 2.58

```{r}
qnorm(0.005, lower.tail=FALSE)
```

]

---

### Confidence interval

.large[
```
             Estimate Std. Error z value Pr(>|z|)    
(Intercept) -2.901344   0.606038  -4.787 1.69e-06
gre          0.003582   0.000986   3.633  0.00028
```

95% confidence interval for $\beta_1$:
]

---

### Motivating the confidence interval

.large[
* $\widehat{\beta}_1$ is our "best guess" of $\beta_1$
* But, there is variability in our estimate of $\beta_1$; $\widehat{\beta}_1$ will vary from sample to sample

**Idea:** What if we take a "new" sample of data, and see how $\widehat{\beta}_1$ changes?
]

---

### Bootstrap sample

.large[
A **bootstrap sample** is a sample of rows from the data, *with* replacement
]

.pull-left[
Original data:

| GRE | Admit |
| --- | --- | 
| 380 | 0 | 
| 660 | 1 |
| 800 | 1 |
| 520 | 0 |
| 760 | 1 |
| 640 | 1 |
]

.pull-left[
Bootstrap sample:

| GRE | Admit |
| --- | --- | 
| 380 | 0 | 
| 380 | 0 |
| 800 | 1 |
| 520 | 0 |
| 760 | 1 |
| 760 | 1 |
]

.large[
**Motivation:** Variability in resampling from the *observed* data should reflect variability in resampling from the *population*
]

---

### Bootstrap sample

.large[
A **bootstrap sample** is a sample of rows from the data, *with* replacement
]

.pull-left[
Original data:

| GRE | Admit |
| --- | --- | 
| 380 | 0 | 
| 660 | 1 |
| 800 | 1 |
| 520 | 0 |
| 760 | 1 |
| 640 | 1 |
]

.pull-left[
Another bootstrap sample:

| GRE | Admit |
| --- | --- | 
| 380 | 0 | 
| 760 | 1 |
| 520 | 0 |
| 800 | 1 |
| 520 | 0 |
| 800 | 1 |
]

.large[
**Motivation:** Variability in resampling from the *observed* data should reflect variability in resampling from the *population*
]

---

### Activity

.large[
* Take many bootstrap samples
* For each bootstrap sample, calculate $\widehat{\beta}_1$
* Examine the distribution of estimated slopes

[https://sta214-f25.github.io/class_activities/ca_15.html](https://sta214-f25.github.io/class_activities/ca_15.html)
]

---

### Activity

.large[
```{r}
nsim <- 1000
boot_ests <- rep(NA, nsim)

for(i in 1:nsim){
  grad_boot <- grad_app |>
    slice_sample(n = nrow(grad_app), replace=T)
  
  m_boot <- glm(admit ~ gre, family = binomial, 
                data = grad_boot)
  boot_ests[i] <- coef(m_boot)[2]
}
```

]

---

### Activity

.large[
```
             Estimate Std. Error z value Pr(>|z|)    
(Intercept) -2.901344   0.606038  -4.787 1.69e-06
gre          0.003582   0.000986   3.633  0.00028
```

```{r}
mean(boot_ests)
sd(boot_ests)
```

.question[
How do the mean and standard deviation of the bootstrap estimates compare to the regression output?
]

]

---

### Activity

```{r, echo=F, fig.width=7, fig.height=5}
hist(boot_ests)
```

.large[
.question[
What does the bootstrap distribution look like?
]
]

---

### Activity

```{r, echo=F, fig.width=7, fig.height=5}
hist(boot_ests)
```

.large[
.question[
Based on the bootstrap distribution, is a value of 0.004 a "reasonable" guess for $\beta_1$?
]
]

---

### Activity

```{r, echo=F, fig.width=7, fig.height=5}
hist(boot_ests)
```

.large[
.question[
Based on the bootstrap distribution, is a value of 0.0025 a "reasonable" guess for $\beta_1$?
]
]

---

### Activity

```{r, echo=F, fig.width=7, fig.height=5}
hist(boot_ests)
```

.large[
.question[
Based on the bootstrap distribution, is a value of 0.1 a "reasonable" guess for $\beta_1$?
]
]

---

### Activity

```{r, echo=F, fig.width=7, fig.height=5}
hist(boot_ests)
```

.large[
.question[
Based on the bootstrap distribution, what would be a "reasonable range" of values for $\beta_1$?
]
]

---

### Confidence interval

.large[
95% confidence interval for $\beta_1$: 

$$\widehat{\beta}_1 \pm 1.96 SE(\widehat{\beta}_1) = 0.00358 \pm 1.96(0.00099) = (0.0016, 0.0055)$$
]

```{r, echo=F, fig.width=7, fig.height=5}
hist(boot_ests)
```

---

### Activity

.large[
Work on the activity (handout).
]

---

### Activity

.large[
```
Coefficients:
            Estimate Std. Error z value Pr(>|z|)    
(Intercept)  1.73743    0.08499   20.44   <2e-16 ***
WBC         -0.36085    0.01243  -29.03   <2e-16 ***
```

.question[
Calculate a 99% confidence interval for the change in the log odds of dengue associated with a one-unit increase in WBC.
]
]

---

### Activity

.large[
```
Coefficients:
            Estimate Std. Error z value Pr(>|z|)    
(Intercept)  1.73743    0.08499   20.44   <2e-16 ***
WBC         -0.36085    0.01243  -29.03   <2e-16 ***
```

.question[
How could we calculate a 99% confidence interval for the change in the *odds* of dengue associated with a one-unit increase in WBC?
]
]





