---
title: Linear mixed effects models
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Data: flipped classrooms?

.large[
Data set has 375 rows (one per student), and the following variables:

* `professor`: which professor taught the class (1 -- 15)
* `style`: which teaching style the professor used (no flip, some flip, fully flipped)
* `score`: the student's score on the final exam
]

---

### Interpretation

.large[
\begin{align}
Score_{ij} &= \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + u_i + \varepsilon_{ij}
\end{align}

$\varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2) \hspace{1cm} u_i \overset{iid}{\sim} N(0, \sigma_u^2)$
]

.large[
* $\widehat{\beta}_0 = 77.66$, $\hspace{1cm} \widehat{\beta}_1 = 11.07$, $\hspace{1cm} \widehat{\beta}_2 = 2.81$
* $\widehat{\sigma}_\varepsilon^2 = 4.25$, $\hspace{1cm} \widehat{\sigma}_u^2 = 21.37$

.question[
How do I interpret $\widehat{\beta}_1 = 11.07$ and $\widehat{\beta}_2 = 2.81$?
]
]

.large[
We expect that, on average, scores in some-flipped classes are 11.07 points higher than for no-flipped classes.

We expect that, on average, scores in fully-flipped classes are 2.81 points higher than for no-flipped classes.
]

---

### Interpretation

.large[
\begin{align}
Score_{ij} &= \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + u_i + \varepsilon_{ij}
\end{align}

$\varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2) \hspace{1cm} u_i \overset{iid}{\sim} N(0, \sigma_u^2)$
]

.large[
* $\widehat{\beta}_0 = 77.66$, $\hspace{1cm} \widehat{\beta}_1 = 11.07$, $\hspace{1cm} \widehat{\beta}_2 = 2.81$
* $\widehat{\sigma}_\varepsilon^2 = 4.25$, $\hspace{1cm} \widehat{\sigma}_u^2 = 21.37$

.question[
How do I interpret $\hspace{1cm} \widehat{\sigma}_u^2 = 21.37$?
]
]

---

### Interpretation

.large[
\begin{align}
Score_{ij} &= \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + u_i + \varepsilon_{ij}
\end{align}

$\varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2) \hspace{1cm} u_i \overset{iid}{\sim} N(0, \sigma_u^2)$
]

.large[
* $\widehat{\beta}_0 = 77.66$, $\hspace{1cm} \widehat{\beta}_1 = 11.07$, $\hspace{1cm} \widehat{\beta}_2 = 2.81$
* $\widehat{\sigma}_\varepsilon^2 = 4.25$, $\hspace{1cm} \widehat{\sigma}_u^2 = 21.37$

.question[
How do I interpret $\hspace{1cm} \widehat{\sigma}_u^2 = 21.37$?
]
]

.large[
The average score varies from professor to professor by about 4.62 ( $= \sqrt{21.37}$ ) points
]

---

### Interpretation

.large[
\begin{align}
Score_{ij} &= \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + u_i + \varepsilon_{ij}
\end{align}

$\varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2) \hspace{1cm} u_i \overset{iid}{\sim} N(0, \sigma_u^2)$
]

.large[
* $\widehat{\beta}_0 = 77.66$, $\hspace{1cm} \widehat{\beta}_1 = 11.07$, $\hspace{1cm} \widehat{\beta}_2 = 2.81$
* $\widehat{\sigma}_\varepsilon^2 = 4.25$, $\hspace{1cm} \widehat{\sigma}_u^2 = 21.37$

.question[
How do I interpret $\widehat{\sigma}_\varepsilon^2 = 4.25$?
]
]

---

### Interpretation

.large[
\begin{align}
Score_{ij} &= \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + u_i + \varepsilon_{ij}
\end{align}

$\varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2) \hspace{1cm} u_i \overset{iid}{\sim} N(0, \sigma_u^2)$
]

.large[
* $\widehat{\beta}_0 = 77.66$, $\hspace{1cm} \widehat{\beta}_1 = 11.07$, $\hspace{1cm} \widehat{\beta}_2 = 2.81$
* $\widehat{\sigma}_\varepsilon^2 = 4.25$, $\hspace{1cm} \widehat{\sigma}_u^2 = 21.37$

.question[
How do I interpret $\widehat{\sigma}_\varepsilon^2 = 4.25$?
]
]

.large[
Within a class, students' scores vary by about  2.06 ( $= \sqrt{4.25}$ ) points
]

---

### Intra-class correlation

.pull-left[
```{r echo=F, message=F, warning=F, fig.width=7, fig.height=4, fig.align='center'}

library(latex2exp)
library(tidyverse)

ann_text_1 <- data.frame(score = 80, professor = 5.5,
                         style = factor("no flip",
                                        levels = c("no flip", "some flip", "full flip")))

ann_text_2 <- data.frame(score = 80, professor = 5.5,
                         style = factor("some flip",
                                        levels = c("no flip", "some flip", "full flip")))

ann_text_3 <- data.frame(score = 80, professor = 5.5,
                         style = factor("full flip",
                                        levels = c("no flip", "some flip", "full flip")))

set.seed(3)

mean_scores <- c(84, 78, 76, 82, 83,
                     82, 80, 90, 91, 87,
                     79, 80, 83, 82, 85)

data.frame(professor = rep(1:15, each=25),
           style = factor(rep(c("no flip", "some flip", "full flip"),
                                 each=125),
                             levels = c("no flip", "some flip", "full flip"))) %>%
  mutate(score = rnorm(375, mean = mean_scores[professor], sd = 6),
         professor = rep(rep(1:5, each=25), 3)) %>%
  ggplot(aes(x = as.factor(professor), y = score)) +
  geom_boxplot(lwd=1) +
  facet_wrap(~style) +
  scale_y_continuous(breaks = c(80, 83, 87),
                     labels = c(TeX("$\\beta_0"),
                                TeX("$\\beta_0 + \\beta_2"),
                                TeX("$\\beta_0 + \\beta_1"))) +
  labs(x = "Professor", y = "Score") +
  theme_classic() +
  theme(text = element_text(size = 30),
        axis.text.x=element_blank(),
        axis.text.y = element_text(size=20)) +
  geom_segment(aes(x = 0.5, y = 80,
                   xend = 5.5, yend = 80),
               data = ann_text_1,
               lwd = 1, lty=3) +
  geom_segment(aes(x = 0.5, y = 87,
                   xend = 5.5, yend = 87),
               data = ann_text_2,
               lwd = 1, lty=3) +
  geom_segment(aes(x = 0.5, y = 83,
                   xend = 5.5, yend = 83),
               data = ann_text_3,
               lwd = 1, lty=3)
```

.large[
$\sigma_\varepsilon^2$ is large relative to $\sigma_u^2$
]
]

.pull-right[
```{r echo=F, message=F, warning=F, fig.width=7, fig.height=4, fig.align='center'}

ann_text_1 <- data.frame(score = 80, professor = 5.5,
                         style = factor("no flip",
                                        levels = c("no flip", "some flip", "full flip")))

ann_text_2 <- data.frame(score = 80, professor = 5.5,
                         style = factor("some flip",
                                        levels = c("no flip", "some flip", "full flip")))

ann_text_3 <- data.frame(score = 80, professor = 5.5,
                         style = factor("full flip",
                                        levels = c("no flip", "some flip", "full flip")))

set.seed(3)

mean_scores <- c(84, 78, 76, 82, 83,
                     82, 80, 90, 91, 87,
                     79, 80, 83, 82, 85)

diffs <- mean_scores - rep(c(80, 87, 83), each=5)

mean_scores <- rep(c(80, 87, 83), each=5) + 3*diffs

data.frame(professor = rep(1:15, each=25),
           style = factor(rep(c("no flip", "some flip", "full flip"),
                                 each=125),
                             levels = c("no flip", "some flip", "full flip"))) %>%
  mutate(score = rnorm(375, mean = mean_scores[professor], sd = 2),
         professor = rep(rep(1:5, each=25), 3)) %>%
  ggplot(aes(x = as.factor(professor), y = score)) +
  geom_boxplot(lwd=1) +
  facet_wrap(~style) +
  scale_y_continuous(breaks = c(80, 83, 87),
                     labels = c(TeX("$\\beta_0"),
                                TeX("$\\beta_0 + \\beta_2"),
                                TeX("$\\beta_0 + \\beta_1"))) +
  labs(x = "Professor", y = "Score") +
  theme_classic() +
  theme(text = element_text(size = 30),
        axis.text.x=element_blank(),
        axis.text.y = element_text(size=20)) +
  geom_segment(aes(x = 0.5, y = 80,
                   xend = 5.5, yend = 80),
               data = ann_text_1,
               lwd = 1, lty=3) +
  geom_segment(aes(x = 0.5, y = 87,
                   xend = 5.5, yend = 87),
               data = ann_text_2,
               lwd = 1, lty=3) +
  geom_segment(aes(x = 0.5, y = 83,
                   xend = 5.5, yend = 83),
               data = ann_text_3,
               lwd = 1, lty=3)
```

.large[
$\sigma_\varepsilon^2$ is small relative to $\sigma_u^2$
]
]

.large[
* Observations within a group are *more correlated* when $\sigma_\varepsilon^2$ is small relative to $\sigma_u^2$
* **Intra-class correlation:** $\rho_{group} = \dfrac{\sigma_u^2}{\sigma_u^2 + \sigma_\varepsilon^2} = \dfrac{\text{between group variance}}{\text{total variance}}$
]

---

### Intra-class correlation

.large[
\begin{align}
Score_{ij} &= \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + u_i + \varepsilon_{ij}
\end{align}

$\varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2) \hspace{1cm} u_i \overset{iid}{\sim} N(0, \sigma_u^2)$
]

.large[
* $\widehat{\beta}_0 = 77.66$, $\hspace{1cm} \widehat{\beta}_1 = 11.07$, $\hspace{1cm} \widehat{\beta}_2 = 2.81$
* $\widehat{\sigma}_\varepsilon^2 = 4.25$, $\hspace{1cm} \widehat{\sigma}_u^2 = 21.37$

$$\widehat{\rho}_{group} = \dfrac{21.37}{21.37 + 4.25} = 0.83$$
]

.large[
So 83% of the variation in student's scores can be explained by differences in average scores from class to class (after accounting for teaching style). That's huge!
]

---

### Class activity

.large[
Work on the class activity (handout).
]

---

### Class activity

.large[
.question[
Interpret the estimate fixed effect coefficients $\widehat{\beta}_0$ and $\widehat{\beta}_1$.
]
]

---

### Class activity

.large[
.question[
Interpret the estimate fixed effect coefficients $\widehat{\beta}_0$ and $\widehat{\beta}_1$.
]
]

.large[
On average (across neighborhoods), we expect that the price of rental with 0 overall satisfaction is $27.28.

For a fixed neighborhood, an increase of 1 point in overall satisfaction is associated with an increase of $14.81 in rental price.
]

---

### Class activity

.large[
.question[
Calculate and interpret the estimated intra-class correlation.
]
]

---

### Class activity

.large[
.question[
Calculate and interpret the estimated intra-class correlation.
]
]

.large[
$\widehat{\rho}_{group} = \dfrac{1048}{1048 + 6762} = 0.134$

About 13% of the variability in price can be explained by differences in the average price between neighborhoods (after accounting for overall satisfaction).
]

---

### Assumptions

.large[
$$Price_{ij} = \beta_0 + \beta_1 Satisfaction_{ij} + u_i + \varepsilon_{ij}$$
$u_i \overset{iid}{\sim} N(0, \sigma_u^2) \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$

where $Price_{ij}$ is the price of rental $j$ in neighborhood $i$
]

.large[
.question[
What assumptions are we making in this mixed effects model?
]
]

---

### Assumptions

.large[
$$Price_{ij} = \beta_0 + \beta_1 Satisfaction_{ij} + u_i + \varepsilon_{ij}$$
$u_i \overset{iid}{\sim} N(0, \sigma_u^2) \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$
]

.large[
* **Shape:** 
  * the overall relationship between satisfaction and price is linear
  * The slope is the *same* for each neighborhood
]

---

### Assumptions

.large[
$$Price_{ij} = \beta_0 + \beta_1 Satisfaction_{ij} + u_i + \varepsilon_{ij}$$
$u_i \overset{iid}{\sim} N(0, \sigma_u^2) \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$
]

.large[
* **Independence:** 
  * random effects are independent
  * observations within neighborhoods are independent after accounting for the random effect (i.e., the random effect captures the correlation within neighborhoods)
  * observations from different neighborhoods are independent
]

---

### Assumptions

.large[
$$Price_{ij} = \beta_0 + \beta_1 Satisfaction_{ij} + u_i + \varepsilon_{ij}$$
$u_i \overset{iid}{\sim} N(0, \sigma_u^2) \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$
]

.large[
* **Normality:** Both $u_i \sim N(0, \sigma_u^2)$ and $\varepsilon_{ij} \sim N(0, \sigma_\varepsilon^2)$
* **Constant variance:** 
  * $\varepsilon_{ij}$ has the same variance $\sigma_\varepsilon^2$ regardless of satisfaction or neighborhood
]

---

### Checking assumptions

.large[
$$Price_{ij} = \beta_0 + \beta_1 Satisfaction_{ij} + u_i + \varepsilon_{ij}$$
$u_i \overset{iid}{\sim} N(0, \sigma_u^2) \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$
]

.large[
* **Shape assumption:** 
  * the overall relationship between satisfaction and price is linear
  * The slope is the *same* for each neighborhood
* **Constant variance assumption:** 
  * $\varepsilon_{ij}$ has the same variance $\sigma_\varepsilon^2$ regardless of satisfaction or neighborhood
  
.question[
How do you think we could check the shape and constant variance assumptions?
]
]

---

### Residual plots

.large[
Residuals: $Price_{ij} - \widehat{Price}_{ij}$, where

$$\widehat{Price}_{ij} = \widehat{\beta}_0 + \widehat{\beta}_1 Satisfaction_{ij} + \widehat{u}_i$$

```{r message=F, warning=F, echo=F, fig.align='center', fig.height=5, fig.width=7}
library(lme4)
library(tidyverse)
library(knitr)

bnb <- read.csv("https://raw.githubusercontent.com/proback/BYSH/master/data/airbnb.csv")
m1 <- lmer(price ~ overall_satisfaction + (1 | neighborhood), data = bnb)

bnb %>%
  mutate(resids = residuals(m1),
         preds = predict(m1)) %>%
  ggplot(aes(x = preds, y = resids)) +
  geom_point(size=2) +
  geom_abline(slope = 0, intercept = 0, color="blue", lwd=1.2) +
  labs(x = "Predicted price", 
       y = "Residual") +
  theme_bw() +
  theme(text = element_text(size = 25))
```
]

---

### Residual plots

```{r message=F, warning=F, echo=F, fig.align='center', fig.height=5, fig.width=7}

bnb %>%
  mutate(resids = residuals(m1),
         preds = predict(m1)) %>%
  ggplot(aes(x = preds, y = resids)) +
  geom_point(size=2) +
  geom_abline(slope = 0, intercept = 0, color="blue", lwd=1.2) +
  labs(x = "Predicted price", 
       y = "Residual") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
.question[
Do the shape and constant variance assumptions look reasonable?
]
]

---


### Residual plots

```{r message=F, warning=F, echo=F, fig.align='center', fig.height=4, fig.width=6}

bnb %>%
  mutate(resids = residuals(m1),
         preds = predict(m1)) %>%
  ggplot(aes(x = preds, y = resids)) +
  geom_point(size=2) +
  geom_abline(slope = 0, intercept = 0, color="blue", lwd=1.2) +
  labs(x = "Predicted price", 
       y = "Residual") +
  theme_bw() +
  theme(text = element_text(size = 25))
```
.large[
* Linearity seems reasonable
* We can't assess (yet) whether the slope is actually the same for all neighborhoods
* Constant variance may be slightly violated (but probably not enough to matter)
]

---

### Calculating predicted values

.large[
Residuals: $Price_{ij} - \widehat{Price}_{ij}$, where

$$\widehat{Price}_{ij} = \widehat{\beta}_0 + \widehat{\beta}_1 Satisfaction_{ij} + \widehat{u}_i$$

.question[
Where do the $\widehat{u}_i$ come from?
]
]

---

### Estimated random effects

.large[
R calculates an estimated random effect for each group (i.e., neighborhood):

```{r}
m1 <- lmer(price ~ overall_satisfaction + 
             (1 | neighborhood), 
           data = bnb)
coef(m1)
```
]

---

### Estimated random effects

.large[

```{r, eval=F}
coef(m1)
```

```
                   (Intercept) overall_satisfaction
Albany Park          16.367331             14.80912
Archer Heights        9.863461             14.80912
Avondale             40.533851             14.80912
Beverly              21.046464             14.80912
Bridgeport           13.304517             14.80912
```

.question[
What is the same for every neighborhood?
]
]

---

### Estimated random effects

.large[

```{r, eval=F}
coef(m1)
```

```
                   (Intercept) overall_satisfaction
Albany Park          16.367331             14.80912
Archer Heights        9.863461             14.80912
Avondale             40.533851             14.80912
Beverly              21.046464             14.80912
Bridgeport           13.304517             14.80912
```

.question[
What is the same for every neighborhood?
]
]

.large[
The slope ( $\widehat{\beta}_1$ )
]

---

### Estimated random effects

.large[

```{r, eval=F}
coef(m1)
```

```
                   (Intercept) overall_satisfaction
Albany Park          16.367331             14.80912
Archer Heights        9.863461             14.80912
Avondale             40.533851             14.80912
Beverly              21.046464             14.80912
Bridgeport           13.304517             14.80912
```

.question[
What is *different* for each neighborhood?
]
]



---

### Estimated random effects

.large[

```{r, eval=F}
coef(m1)
```

```
                   (Intercept) overall_satisfaction
Albany Park          16.367331             14.80912
Archer Heights        9.863461             14.80912
Avondale             40.533851             14.80912
Beverly              21.046464             14.80912
Bridgeport           13.304517             14.80912
```

.question[
What is *different* for each neighborhood?
]
]

.large[
The intercept ( $\widehat{\beta}_0 + \widehat{u}_i$ )
]

---

### Estimated random effects

.large[

```
                   (Intercept) overall_satisfaction
Albany Park          16.367331             14.80912
Archer Heights        9.863461             14.80912
Avondale             40.533851             14.80912
Beverly              21.046464             14.80912
Bridgeport           13.304517             14.80912
```

.question[
How do I get the random effect estimates $\widehat{u}_i$ ?
]
]

---

### Estimated random effects

.large[

```
                   (Intercept) overall_satisfaction
Albany Park          16.367331             14.80912
Archer Heights        9.863461             14.80912
Avondale             40.533851             14.80912
Beverly              21.046464             14.80912
Bridgeport           13.304517             14.80912
```

.question[
How do I get the random effect estimates $\widehat{u}_i$ ?
]
]

.large[
The intercept for each neighborhood is $\widehat{\beta}_0 + \widehat{u}_i$

So subtract the estimate $\widehat{\beta}_0$ from the intercept for each neighborhood
]

---

### Checking assumptions

.large[
$$Price_{ij} = \beta_0 + \beta_1 Satisfaction_{ij} + u_i + \varepsilon_{ij}$$
$u_i \overset{iid}{\sim} N(0, \sigma_u^2) \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$
]

.large[
* **Normality assumption:** Both $u_i \sim N(0, \sigma_u^2)$ and $\varepsilon_{ij} \sim N(0, \sigma_\varepsilon^2)$
  
.question[
How do you think we could check the normality assumption?
]
]

---

### QQ plots

.large[
**Assumption:** $u_i \sim N(0, \sigma_u^2)$

* Check whether the random effect estimates $\widehat{u}_i$ appear normal with a QQ plot

**Assumption:** $\varepsilon_{ij} \sim N(0, \sigma_\varepsilon^2)$

* Check whether the residuals appear normal with a QQ plot
]

---

### QQ plot for the residuals

```{r message=F, warning=F, echo=F, fig.align='center', fig.height=5, fig.width=7}

bnb %>%
  mutate(resids = residuals(m1)) %>%
  ggplot(aes(sample = resids)) +
  geom_qq(size=2) +
  geom_qq_line(color="blue", lwd=1.2) +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed residual quantiles") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
.question[
Do the residuals appear normal?
]
]

---

### QQ plot for the residuals

```{r message=F, warning=F, echo=F, fig.align='center', fig.height=5, fig.width=7}

bnb %>%
  mutate(resids = residuals(m1)) %>%
  ggplot(aes(sample = resids)) +
  geom_qq(size=2) +
  geom_qq_line(color="blue", lwd=1.2) +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed residual quantiles") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
.question[
Do the residuals appear normal?
]
]

.large[
No -- though the sample size is large, so it might not matter
]

---

### QQ plot for the random effects

```{r message=F, warning=F, echo=F, fig.align='center', fig.height=5, fig.width=7}

data.frame(re = coef(m1)$neighborhood[,1] - summary(m1)$coefficients[2,1]) %>%
  ggplot(aes(sample = re)) +
  geom_qq(size=2) +
  geom_qq_line(color="blue", lwd=1.2) +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed random effect quantiles") +
  theme_bw() +
  theme(text = element_text(size = 20))
```

.large[
.question[
Do the random effects appear normal?
]
]

---

### QQ plot for the random effects

```{r message=F, warning=F, echo=F, fig.align='center', fig.height=5, fig.width=7}

data.frame(re = coef(m1)$neighborhood[,1] - summary(m1)$coefficients[2,1]) %>%
  ggplot(aes(sample = re)) +
  geom_qq(size=2) +
  geom_qq_line(color="blue", lwd=1.2) +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed random effect quantiles") +
  theme_bw() +
  theme(text = element_text(size = 20))
```

.large[
.question[
Do the random effects appear normal?
]
]

.large[
No -- and this could be more problematic
]

---

### Addressing assumption violations


.pull-left[
```{r message=F, warning=F, echo=F, fig.align='center', fig.height=4, fig.width=6}

bnb %>%
  mutate(resids = residuals(m1)) %>%
  ggplot(aes(sample = resids)) +
  geom_qq(size=2) +
  geom_qq_line(color="blue", lwd=1.2) +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed residual quantiles") +
  theme_bw() +
  theme(text = element_text(size = 20))
```
]

.pull-right[
```{r message=F, warning=F, echo=F, fig.align='center', fig.height=4, fig.width=6}

data.frame(re = coef(m1)$neighborhood[,1] - summary(m1)$coefficients[2,1]) %>%
  ggplot(aes(sample = re)) +
  geom_qq(size=2) +
  geom_qq_line(color="blue", lwd=1.2) +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed random effect quantiles") +
  theme_bw() +
  theme(text = element_text(size = 20))
```
]

.large[
.question[
How could we address violations of the normality assumptions?
]
]

---

### Addressing assumption violations


.pull-left[
```{r message=F, warning=F, echo=F, fig.align='center', fig.height=4, fig.width=6}

bnb %>%
  mutate(resids = residuals(m1)) %>%
  ggplot(aes(sample = resids)) +
  geom_qq(size=2) +
  geom_qq_line(color="blue", lwd=1.2) +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed residual quantiles") +
  theme_bw() +
  theme(text = element_text(size = 20))
```
]

.pull-right[
```{r message=F, warning=F, echo=F, fig.align='center', fig.height=4, fig.width=6}

data.frame(re = coef(m1)$neighborhood[,1] - summary(m1)$coefficients[2,1]) %>%
  ggplot(aes(sample = re)) +
  geom_qq(size=2) +
  geom_qq_line(color="blue", lwd=1.2) +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed random effect quantiles") +
  theme_bw() +
  theme(text = element_text(size = 20))
```
]

.large[
.question[
How could we address violations of the normality assumptions?
]
]

.large[
Transformations!
]

---

### Transformations

.large[
$$\log(Price_{ij}) = \beta_0 + \beta_1 Satisfaction_{ij} + u_i + \varepsilon_{ij}$$
$u_i \overset{iid}{\sim} N(0, \sigma_u^2) \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$
]

```{r include =F}
m1 <- lmer(log(price) ~ overall_satisfaction + (1 | neighborhood), data = bnb)
```

.pull-left[
```{r message=F, warning=F, echo=F, fig.align='center', fig.height=4, fig.width=6}

bnb %>%
  mutate(resids = residuals(m1)) %>%
  ggplot(aes(sample = resids)) +
  geom_qq(size=2) +
  geom_qq_line(color="blue", lwd=1.2) +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed residual quantiles") +
  theme_bw() +
  theme(text = element_text(size = 20))
```
]

.pull-right[
```{r message=F, warning=F, echo=F, fig.align='center', fig.height=4, fig.width=6}

data.frame(re = coef(m1)$neighborhood[,1] - summary(m1)$coefficients[2,1]) %>%
  ggplot(aes(sample = re)) +
  geom_qq(size=2) +
  geom_qq_line(color="blue", lwd=1.2) +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed random effect quantiles") +
  theme_bw() +
  theme(text = element_text(size = 20))
```
]

.large[
* Transformations can help with both the residuals and random effects
]








