---
title: Inference with mixed effects
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Agenda

.large[
* Project part 3 due last day of class
  * No additional analysis needed; just combining work from first two parts and adding introduction, conclusion, etc.
  * There will be time during the last week of class to work on the project
* Exam 2 Friday
  * Review day on Wednesday in class
]

---

### Data and Goal

.large[
```{r include=F}
library(lme4)
library(tidyverse)
library(knitr)

bryozoan <- read_csv("bryozoan_data.csv")

bugula_early <- bryozoan %>%
  filter(Stage == "early", Species == "bugula")
```

We have date on 197 early-stage *Bugula neritina*, with information on

* `Run`: which repetition of the experiment the individual was recorded in
* `Mass`: Mass of the individual (in micrograms)
* `Metabolic`: Recorded metabolic rate (rate of energy consumption) of the individual (in mJ per hour)

.question[
**Goal for this class:** Is there systematic variation between different runs (i.e., is there any correlation due to Run)?
]
]

---

### Visualizing runs

.large[

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
bugula_early %>%
  ggplot(aes(x = as.factor(Run), y = Metabolic)) +
  geom_boxplot() +
  theme_bw() +
  labs(x = "Run") +
  theme(text = element_text(size = 15))
```

.question[
Does it look like there is variation from run to run (i.e., do we need $u_i$ )?
]
]

---

### Visualizing runs

.large[

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
bugula_early %>%
  ggplot(aes(x = as.factor(Run), y = Metabolic)) +
  geom_boxplot() +
  theme_bw() +
  labs(x = "Run") +
  theme(text = element_text(size = 15))
```

It is hard to tell if there are actually differences between runs, or if observed differences are just due to chance.

.question[
How can I estimate variability between runs?
]
]

---

### Initial random intercepts model

.large[
$$Metabolic_{ij} = \beta_0 + u_i + \varepsilon_{ij} \hspace{1cm} u_i \overset{iid}{\sim} N(0, \sigma_u^2), \hspace{0.2cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$$

```{r eval=F}
m1 <- lmer(Metabolic ~ (1|Run), data = bugula_early)
summary(m1)
```

```
Random effects:
 Groups   Name        Variance Std.Dev.
 Run      (Intercept) 0.000131 0.01145 
 Residual             0.004181 0.06466
```
]


---

### Initial random intercepts model

.large[
$$Metabolic_{ij} = \beta_0 + u_i + \varepsilon_{ij} \hspace{1cm} u_i \overset{iid}{\sim} N(0, \sigma_u^2), \hspace{0.2cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$$

$\widehat{\beta}_0 = 0.175$, $\hspace{0.5cm} \widehat{\sigma}_u^2 = 0.00013$, $\hspace{0.5cm} \widehat{\sigma}_\varepsilon^2 = 0.0042$

.question[
How would we test whether there is systematic variation between runs?
]

]

---

### Plan

---

### Class activity

.large[
Work on the class activity. Submit your HTML on canvas at the end of class.
]