---
title: "Class activity: bootstrap sampling"
format: html
---

**Instructions:** 

* Work individually or with a neighbor to answer the following questions 
* To get started, download the [class activity template](https://sta214-f25.github.io/class_activities/ca_15_template.qmd) file
* When you are finished, render the file as an HTML and submit the HTML to Canvas (let me know if you encounter any problems)

## Bootstrap sampling

A **bootstrap sample** is a sample of rows from the original data, *with replacement*. Bootstrap sampling can be used to explore the sampling distribution and variability of a statistic. The idea is that variability in resampling from the *observed* data should reflect variability in resampling from the *population*.

### Grad app data

Suppose we are working with the grad app data, and the model

$$Admit_i \sim Bernoulli(\pi_i)$$

$$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 GRE_i$$

We fit the model in R, and we are interested in estimating $\beta_1$:

```
             Estimate Std. Error z value Pr(>|z|)    
(Intercept) -2.901344   0.606038  -4.787 1.69e-06
gre          0.003582   0.000986   3.633  0.00028
```

Now we would like to explore how $\widehat{\beta}_1$ varies from sample to sample.

The code below creates a bootstrap sample of the data, and re-fits the logistic regression model on the bootstrap sample:

```{r, eval=F}
library(tidyverse)

# bootstrap sample
grad_boot <- grad_app |>
  slice_sample(n = nrow(grad_app), replace=T)

# re-fit model
m_boot <- glm(admit ~ gre, family = binomial, data = grad_boot)

# get estimated slope
coef(m_boot)[2]
```

1. Run the code; how does the estimate of $\beta_1$ from the bootstrap sample compare to the estimate from the original data?

<br/>

Now let's repeat the process many times, using a `for` loop! Here is some starter code:

```{r, eval=F}
nsim <- 1000
boot_ests <- rep(NA, nsim) # vector to store slope estimates

for(i in 1:nsim){
  # take a bootstrap sample
  # fit the model
  # store the estimated slope (fill in the ...)
  boot_ests[i] <- ...
}
```

2. Fill in the code to estimate $\beta_1$ in 1000 bootstrap samples.

3. Calculate the mean of the bootstrap estimates. How does the mean compare to $\widehat{\beta}_1$ from the original data?

4. Calculate the standard deviation of the bootstrap estimates. How does the standard deviation compare to the reported standard error $SE(\widehat{\beta}_1)$ from the logistic regression output on the original data?

5. Make a histogram of the bootstrap estimates, and describe the distribution. Based on this bootstrap distribution, provide a "reasonable range" of values for 
$\beta_1$.



