---
title: "Activity"
format: pdf
editor: source
---

\vspace{-2.5cm}

**Group members:**

## Grap applications data

Suppose we are interesting in testing whether there is a relationship between college GPA and acceptance to a graduate program. We fit the regression model

$$Admit_i \sim Bernoulli(\pi_i)$$

$$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 GPA_i$$

and we wish to test the hypotheses $H_0: \beta_1 = 0$ vs. $H_A: \beta_1 \neq 0$. We will use the drop-in-deviance statistic

$$G = \text{deviance}_{\text{reduced}} - \text{deviance}_{\text{full}}$$

Here is model output for the full and reduced models, and a histogram approximating the null distribution for the drop-in-deviance statistic $G$:

```{r}
grad_app <- read.csv("https://stats.idre.ucla.edu/stat/data/binary.csv")

# full model
glm(admit ~ gpa, family = binomial, data = grad_app)
# reduced model
glm(admit ~ 1, family = binomial, data = grad_app)
```

```{r, echo=F}
hist(rchisq(1000, df=1), xlab="Drop-in-deviance (G)", 
     main = "Approximate null distribution")
```


1. What do you notice about the "Null Deviance" for the two models? How does it compare to the "Residual Deviance"?

\hspace{3cm}

2. Calculate the drop-in-deviance statistic for the hypotheses above.

\hspace{3cm}

3. Based on the observed drop-in-deviance and the null distribution, do you think that the null hypothesis is true?







