---
title: "Class activity: model selection"
format: html
---

## SBA data

In this activity, we will experiment with stepwise selection, using AIC to compare models on the SBA data. Recall the SBA data contains the following variables:

* `Default`: whether the business defaulted on the loan (1 = yes, 0 = no)
* `UrbanRural`: 1 if business is in urban area, 2 if business is in rural area, 0 if unknown
* `NewExist`: 1 if business already existed, 2 if business is new, 0 if unknown
* `Term`: Length of the loan term (months)
* `NoEmp`: Number of employees of the business before receiving the loan
* `GrAppv`: Gross amount of loan approved by the bank, in dollars
* `SBA_Appv`: Amount of the loan guaranteed by the SBA, in dollars
* `DisbursementGross`: The amount of money disbursed (loaned), in dollars

You can load the SBA data into R with

```{r}
sba <- read.csv("https://sta214-f25.github.io/class_activities/sba_small.csv")
```

## Questions

The code below uses forward stepwise selection to choose a model on the SBA data. The smallest model considered (our starting model) is the intercept-only model, and the largest model considered (`scope` in the code) includes all the available variables. The code uses the `stepAIC` function in the `MASS` package, which you may need to install.

```{r, message=F, warning=F}
library(MASS)
library(tidyverse)

# specify the starting model (intercept-only)
m0 <- glm(Default ~ 1, data = sba, family = binomial)

# forward selection using AIC
# Note we have to specify the largest model we want to consider
forward_aic <- stepAIC(m0, scope = ~ Term + DisbursementGross + 
                         factor(UrbanRural) + NoEmp +
                         GrAppv + SBA_Appv,
                       direction = "forward",
                       trace = 0)

summary(forward_aic)
```

1. Run the code. Which variables are chosen using forward selection with AIC?

**Solution:** Term, Urban/Rural, number of employees, gross approval, gross disbursement

2. Think back to our previous class activity, in which we performed model diagnostics with the SBA data. Are there any issues with our final fitted model? How would we correct them?

**Solution:** There are several issues with the fitted model. First, we know from prior exploration with the SBA data that transformations (e.g., a log transform) are needed on several variables, such as gross disbursement. Second, we know that gross disbursement and gross approval are highly correlated. The issue is that we have not specified either of this in our stepwise selection approach above. To address these issues, we would need to change the model that the stepwise procedure is allowed to consider -- by transforming some variables, and removing other variables from consideration.

3. Will stepwise selection detect multicollinearity in the explanatory variables?

**Solution:** No -- stepwise selection doesn't check for multicollinearity, and is perfectly happy to add correlated variables to the model as long as the AIC goes down. We can see the result of this above!

4. Will stepwise selection fix violations to the shape assumption?

**Solution:** No -- stepwise selection here isn't considering any transformations on our variables, nor is it checking whether the shape assumption is reasonable.

5. Using transformations to fix violations to the shape assumption, and choosing a subset of variables to handle issues with multicollinearity, modify the `scope` of the forward stepwise procedure. Which variables are now selected by the procedure?

**Solution:**

```{r}
forward_aic <- stepAIC(m0, scope = ~ Term + log(DisbursementGross) + 
                         factor(UrbanRural) + NoEmp,
                       direction = "forward",
                       trace = 0)

summary(forward_aic)
```



