---
title: "Class activity: Poisson regression assumptions"
format: html
---

**Instructions:** 

* Work individually or with a neighbor to answer the following questions 
* To get started, download the [class activity template](https://sta214-f25.github.io/class_activities/ca_21_template.qmd) file
* When you are finished, render the file as an HTML and submit the HTML to Canvas (let me know if you encounter any problems)

# Poisson distribution assumptions

Consider the Poisson regression model

$$Y_i \sim Poisson(\lambda_i)$$

$$\log(\lambda_i) = \beta_0 + \beta_1 X_i$$

As we discussed in class, this model makes assumptions about shape, distribution, independence, and randomness. As always, to assess the independence and randomness assumptions we have to think about the data generating process. For the shape and distribution assumptions, however, we can use diagnostic plots.

There are a variety of ways to assess these assumptions. In this class activity, we will look at two visual tools: **quantile residual plots** and **rootograms**. Quantile residual plots can provide information about both the distribution assumption and the shape assumption; rootograms provide information about how well our model fits the observed data.


# Example

To begin, let's generate data from the model

$$Y_i \sim Poisson(\lambda_i)$$
$$\log(\lambda_i) = X_i$$

where $X_i \sim N(0, 0.5)$. Note that the Poisson regression assumptions are satisfied for this model.

```{r}
n <- 1000
x <- rnorm(n, sd = 0.5)
y1 <- rpois(n, lambda = exp(x))

m1 <- glm(y1 ~ x, family = poisson)
summary(m1)
```

## Quantile residual plot

Using the `qresid` function from the `statmod` package, we can construct a quantile residual plot for our fitted model:

```{r, message=F, warning=F}
library(statmod)
library(tidyverse)

data.frame(x = x, resids = qresid(m1)) |>
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(x = "X", y = "Quantile residuals") 
```

Interpreting quantile residual plots is similar to interpreting residual plots in a linear regression model. Some features to note:

* The residuals are randomly scattered around 0, with no clear pattern
* The variance of the residuals appears constant
* Most of the residuals lie between -2 and 2 (if the model is a good description of the data, the quantile residuals should be approximately standard normal)


## Rootogram
 
A **rootogram** is a visual tool which compares a histogram of the observed counts for the response variable (on the square root scale) to the predicted frequencies from the fitted model:

```{r}
library(topmodels)
rootogram(m1, style="standing")
```

There is one bar for each distinct value of the response variable, and the height of the bar is the square root of the frequency of that value in the observed data. The red points and lines represent the *predicted* frequency of each value, using the fitted model. The dashed horizontal lines at each bar provide a margin of error for the estimates.

*If the model is a good fit to the data*:

* The histogram bars will be close to the predictions (in red) from the fitted model

# Breaking the Poisson distribution assumption

The data we generated above really do come from a Poisson distribution (we used the `rpois` function to simulate it!). What happens if the assumption of Poisson data is *wrong*?

The code below generates data where the response comes from a distribution called the Negative Binomial, which is *not* a Poisson. However, we still fit a Poisson regression model, so we are using the wrong distribution in our model:

```{r, eval=F}
n <- 1000
x <- rnorm(n, sd = 0.5)
y2 <- rnbinom(n, size=0.5, mu=exp(x))

m2 <- glm(y2 ~ x, family = poisson)
```

1. Re-create the quantile residual plot. What has changed in the quantile residual plot? How does it compare to the plot when all the assumptions were satisfied?

2. Re-create the rootogram. What has changed? How does it compare to the plot when all the assumptions were satisfied?


# Breaking the shape assumption

So far, we have seen what happens if the shape assumption is correct, but the Poisson distribution assumption is wrong. What happens if the Poisson distribution assumption is correct, but the *shape* assumption is wrong?

The code below simulates data for which the shape assumption is wrong:

```{r, eval=F}
n <- 1000
x <- rnorm(n, sd = 0.5)
y3 <- rpois(n, lambda = exp(x^2))

m3 <- glm(y3 ~ x, family = poisson)
```

3. Re-create the quantile residual plot. What has changed in the quantile residual plot? How does it compare to the plot when all the assumptions were satisfied?

4. Re-create the rootogram. What has changed? How does it compare to the plot when all the assumptions were satisfied?




