---
title: "Class activity"
format: html
---

**Instructions:** There is nothing you need to submit for this class activity. Play around with the simulation, and see how each parameter influences the resulting data!

# Simulating data from a mixed effects model

The code below simulates data from a mixed effects model, and creates boxplots showing the distribution:

```{r, eval=F}
library(tidyverse)

# simulate flipped classroom data
# 15 classrooms (professors), 5 from each teaching style,
# with 25 students per class
flipped_data <- data.frame(professor = rep(1:15, each=25),
                           style = factor(rep(c("no flip", 
                                                "some flip",
                                                "full flip"), 
                                              each=125), 
                                          levels = c("no flip", 
                                                     "some flip", 
                                                     "full flip")))

# true parameters for the underlying mixed effects model                          
beta0 <- 80
beta1 <- 5
beta2 <- 3
sigma2 <- 2
sigma2_u <- 1

# simulate the random effect terms
# there is one random effect per group (class)
random_effects <- rnorm(15, sd = sqrt(sigma2_u))

# create simulated test scores
flipped_data$test_scores <- beta0 + 
  beta1*(flipped_data$style == "some flip") +
  beta2*(flipped_data$style == "full flip") + 
  rnorm(nrow(flipped_data), sd = sqrt(sigma2)) +
  random_effects[flipped_data$professor]

# plot scores for each class (professor), colored by teaching style
flipped_data |>
  ggplot(aes(x = as.factor(professor), y = test_scores,
             color = style)) +
  geom_boxplot(lwd=1) +
  labs(x = "Professor", y = "Score") +
  theme_classic()
```

1. Run the code, and experiment with changing the parameters. What happens when you increase/decrease $\sigma_u^2$ (`sigma2_u` in the code)? What about when you change $\sigma_\varepsilon^2$ (`sigma2` in the code)? What about $\beta_1$ and $\beta_2$?

2. Now run the code below to fit the mixed effects model. Which pieces of the output correspond to the simulation parameters?
  * `test_scores ~ style + (1|professor)` means that `test_scores` is the response, `style` is a fixed effect, and `professor` is a random effect. In particular, we have a random *intercept* for each professor

```{r, eval=F}
library(lme4)

mixed_mod <- lmer(test_scores ~ style + (1|professor), data = flipped_data)
summary(mixed_mod)
```



