---
title: "Class activity"
format: html
---

**Instructions:** 

* Work individually or with a neighbor to answer the following questions 
* To get started, download the [class activity template](https://sta214-f25.github.io/class_activities/ca_27_template.qmd) file
* When you are finished, render the file as an HTML and submit the HTML to Canvas (let me know if you encounter any problems)

# Diagnostics for ZIP models

In this class activity, we will create quantile residual plots to check the shape assumption for ZIP models

## Questions

1. Run the code below to generate ZIP data for which the shape assumptions are satisfied for both the logistic and the Poisson components.

```{r, eval=F}
x <- rnorm(1000)
alpha <- exp(x)/(1 + exp(x))
lambda <- exp(1 + x)
z <- rbinom(1000, 1, prob=alpha)
y <- 0*z + rpois(1000, lambda)*(1 - z)
```

2. Now use the code below to fit a ZIP model, and create a quantile residual plot for your fitted model. Are the fitted regression coefficients close to the true parameters?

```{r, eval=F}
library(pscl)
library(tidyverse)

## randomized quantile residuals for a ZIP model
## zip_m = fitted ZIP model from pscl
qresid_zeroinfl <- function(zip_m){
  y <- zip_m$y
  pred_probs <- predict(zip_m, type="prob")
  resids <- rep(NA, length(y))
  for(i in seq_along(y)){
    cdf_b <- sum(pred_probs[i,1:(y[i]+1)])
    if(y[i] == 0){
      cdf_a <- 0
    } else {
      cdf_a <- sum(pred_probs[i,1:y[i]])
    }
    
    resids[i] <- qnorm(runif(1, cdf_a, cdf_b))
  }
  return(resids)
}


m <- zeroinfl(y ~ x | x)
summary(m)

data.frame(x = x, resids = qresid_zeroinfl(m)) |>
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  theme_bw()
```

3. Now generate data so that the Poisson or logistic shape assumption is violated. Re-fit the ZIP model, and make a quantile residual plot for the new model. Can you detect the shape violations in the residual plot?

4. If you see a violation in the quantile residual plot, will you be able to tell whether the violation is in the Poisson component or in the logistic component?