---
title: "Final exam review questions"
format: html
---

**The final exam in this class will be cumulative**, covering everything we have learned from logistic regression through mixed effects models. 

To review the material from logistic regression and count regression (Poisson, quasi-Poisson, ZIP, etc.), look back at the review questions for the first two exams.

Below are questions on mixed effects models to help you study this portion of the course. These are some examples of the kinds of questions I might ask.

## Data

Here we will work with data on the mobility scores for 8510 patients seen by doctors in a large healthcare system. Each row in the data represents one patient, and we have the following variables recorded for each patient:

* `mobility`: their mobility score. This is a continuous variable, and higher values represent greater mobility
* `sex`: the patient's sex, recorded as male (sex = 0) or female (sex = 1)
* `age`: the patient's age in years
* `married`: the patient's marital status (1 if married, 0 if not married)
* `doctorID`: and ID for the doctor that the patient visited (values from 1 to 230). Each doctor sees multiple patients in the dataset

Researchers believe that mobility scores will vary from patient to patient, and also from doctor to doctor. Furthermore, they believe that mobility scores depend on the patient's sex, age, and marital status.

## Questions

1. Mixed effects models are useful when there are group effects in our data. What are the groups in this data, and what are the observations within each group?

2. Write down a population mixed effects model that matches the researchers' beliefs. Make sure to include all pieces of the population model, and explain what your subscripts represent!

### Fitted model

We now fit the mixed effects model in R, producing the following output:

```
Random effects:
 Groups   Name        Variance Std.Dev.
 doctorID (Intercept) 0.165    0.4062  
 Residual             1.038    1.0189  
Number of obs: 8510, groups:  doctorID, 230

Fixed effects:
             Estimate Std. Error 
(Intercept) -0.154277   0.148705  
sex          0.522167   0.022378 
age          0.051686   0.002234 
married      0.251665   0.022367 
```


```{r, include=F, eval=F}
set.seed(3)
m <- 230
n <- 37
sex <- rbinom(m*n, 1, 0.5)
age <- rnorm(m*n, 65, 5)
married <- rbinom(m*n, 1, 0.5)
doctorID <- rep(1:m, each=n)
u <- rnorm(m, sd=0.4)
noise <- rnorm(m*n)
mobility <- 0.5*sex + 0.05*age + 0.3*married + u[doctorID] + noise

m1 <- lmer(mobility ~ sex + age + married + (1|doctorID))
summary(m1)
```

3. Interpret each of the estimated fixed effects coefficients in the model.

4. Interpret the values 0.165 and 1.038 in the model output.

5. Using the model output, does there appear to be systematic variation in mobility scores between doctors? Calculate an appropriate statistic.

6. Below is output for the 214th doctor in the dataset. For doctor 214, what is the predicted mobility score for a female patient, unmarried, of age 65?

```
doctorID    (Intercept)       sex        age   married
214          -0.2902217 0.5221674 0.05168561 0.2516646
```

### Inference

7. Suppose you want to test whether there is systematic variation in mobility scores between doctors, after accounting for patient sex, age, and marital status. Write down the null and alternative hypotheses, and describe what your reduced model would be (you may treat the model whose output is shown above as your full model).

8. Now suppose instead that you want to test whether there is a relationship between patients' marital status and mobility scores, after accounting for sex, age, and systematic variation between doctors. Write down the null and alternative hypotheses, and describe what your reduced model would be (you may treat the model whose output is shown above as your full model).

Now you fit your reduced model from question 8, producing the following output:

```
Random effects:
 Groups   Name        Variance Std.Dev.
 doctorID (Intercept) 0.1746   0.4179  
 Residual             1.0538   1.0266  
Number of obs: 8510, groups:  doctorID, 230

Fixed effects:
             Estimate Std. Error 
(Intercept) -0.043115   0.149453 
sex          0.540829   0.022546 
age          0.053930   0.002251
```

9. Describe how you would use parametric bootstrapping to carry out the hypothesis test from the question. Provide as much detail as you can, so that someone could turn your description into R code if they wanted to (you do not need to write code, though you may choose to if it helps you explain your procedure). Your description should include details like values for the parameters of the model you will simulate from, how many simulations you will use, and how you will calculate a test statistic for each simulation.

```{r, include=F, eval=F}
m2 <- lmer(mobility ~ sex + age + (1|doctorID))
summary(m2)
```







